---
title: "2026 Time Series"
author: "05-arima-diagnostics"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
editor_options: 
  chunk_output_type: console
---

::: {.hidden}
$$
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\newcommand\mcy{{\bf y}}
$$
:::

```{r, include=FALSE}
library(forecast)
library(fpp3)
library(tidyverse)
library(tsbox)
library(zoo)
library(seasonal)
library(astsa)
library(patchwork)
library(tseries)
```

```{r setup, include=FALSE}
# Set up chunk for all slides
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  dev.args = list(pointsize = 11)
)
```

# ---- Estimation

## Method of Moments 

The idea behind these estimators is twidewidehat of equating population moments to sample moments and then solving for the parameters in terms of the sample moments.

Start with AR(p) model

$$
x_t = \phi_1 x_{t-1} + \cdots + \phi_p x_{t-p} + w_t
$$

## Yule Walker Estimators
  
From this AR(p) recursion we can derive moment based estimators:

$$
\widehat{\phi} = \widehat{\Gamma}^{-1}\widehat{\gamma}, \quad \widehat{\sigma}^2 = \widehat{\gamma}(0) - \widehat{\gamma}^\prime \widehat{\Gamma}^{-1}\widehat{\gamma}
$$

::: {.callout-tip}
## Exercise for you
Show how to go from AR(p) recursion to the Yule-Walker estimator

Hint: Multiply through by $X_t$ and take expected value
:::

  
## Yule Walker Example
  
```{r}
set.seed(123)
x = arima.sim(model = list(ar = c(1.5, -.75)), 
              n = 144)
sample_acf = acf(x, plot = FALSE)$acf
Gam_hat = toeplitz(sample_acf[1:2])
gam_hat = sample_acf[2:3]
solve(Gam_hat) %*% gam_hat
ar.yw(x)
```

## OLS Estimation of AR(p)

The AR equation is suitable for least squares estimation.

Let ${\bf X}_t = (1 : X_t : \ldots,X_{t-p+1}),$
${\bf X} = [{\bf X}_p': \cdots : {\bf X}_{T-1}']',$  ${\bf Y} = (X_{p+1}, \ldots, X_T)',$
${\bf Z} = (Z_{p+1}, \ldots, Z_T)'$
and $\boldsymbol{\beta} = (\alpha: \phi_1: \ldots \phi_p)'.$ Then the data equation can be written in the form of a multiple linear regression (MLReg)

$$ {\bf Y} = {\bf X}\boldsymbol{\beta} + {\bf Z} $$

The standard estimation for MLReg applies here as well. The **ordinary least squares (OLS)** estimator is 

$$\hat{\boldsymbol{\beta}} = ({\bf X}'{\bf X})^{-1}{\bf X}'{\bf Y}$$

## Gaussian Likelihood

- In this we assume $\{X_t\}$ is a Gaussian time series
  * mean zero
  * ACVF $\gamma(h) = E[X_t X_{t+h}]$
  
- Let ${\bf X}_n = (X_1, \ldots, X_n)^\prime$
- $E[{\bf X}_n {\bf X}_n^\prime] = \Gamma_n$ an $n\times n$ matrix (assume nonsingular)

The likelihood of ${\bf X}_n$ is:

$$
L(\Gamma_n)
=
(2\pi)^{-n/2}
(det ~\Gamma_n)^{-1/2}
exp\left( \frac{1}{2} {\bf X}_n^\prime \Gamma_n^{-1} {\bf X}_n \right)
$$

## Alternative GL representations

- The direct calculation of $\det(\Gamma_n)$ and $\Gamma_n^{-1}$ can be avoided by expressing this in terms of the one-step predictors $\hat{X_i}$, and their mean squared errors $v_{i-1}$
- Both are easily calculated *recursively* from the **innovations algorithm**

## GL representation

For a normal ARMA(p,q) model, let $\beta = (\mu, \phi_1,...,\phi_p, \theta_1,...,\theta_q)'$ be the $(p + q + 1)$-dimensional vector of the model parameters. 

The likelihood can be written as

$$
L(\beta, \sigma_w^2) = {\prod_{t=1}^{n} f(x_t \,|\, x_{t-1},...,x_1)}.
$$

The conditional distribution of $x_t$ given $x_{t-1},...,x_1$ is Gaussian with mean $x_{t-1}$ and variance $P_t$.


## 


The likelihood can be written in terms of one-step ahead conditional densities 
$$ L(\beta, \sigma_z^2|data) = \prod_{t=1}^T N(X_t| X_{t}^{t-1}, P_t^{t-1})$$
w
here $X_t^{t-1} = E(X_t| X_{t-1}, X_{t-2}, \ldots, X_1)$ is the one-step ahead predictor (BLP) for the Gaussian model and 

$$ P_t^{t-1} = \gamma(0)\prod_{j=1}^{t-1}(1 - \phi_{jj}^2)$$


## Parameter estimation 

- In practice, we maximize the GL to find parameter estimates
- Most software will return the value of the log likelihood of the data
  * the logarithm of the probability of the observed data coming from the estimated model
- 

## Information Criteria

Good models are obtained by minimising the AIC, AICc or BIC

- $\text{AIC} = -2 \log(L) + 2(p + q + k + 1)$
- $\text{AICc} = \text{AIC} + \frac{2(p + q + k + 1)(p + q + k + 2)}{T - p - q - k - 2}$
- $\text{BIC} = \text{AIC} + \left[\log(T) - 2\right](p + q + k + 1)$

# ---- Diagnostics

## Residuals

Define residuals

## Residual autocorrelation

- Plotting $\widehat{\rho}_w(h)$ can be a good visual test
- We can perform a general test that take into consideration the magnitudes of the residual ACF lags as a group

Ljung-Box-Pierce Q-statistic:
$$
Q = 
n(n+2)
\sum_{h=1}^{H}
\frac{\widehat{\rho}^2_w(h)}{n-h}
$$

## LBQ stat

- Under the null hypothesis of zero ACF
$$
Q \sim \chi^2_{H-p-q}
$$
- Thus we would reject the null at level $\alpha$ if the value of $Q$ exceeds the $(1-\alpha)$-quantile of the $\chi^2$ distribution. 

## Example Box-Jenkins fitting method

Lets fit an ARIMA model to the following data:

```{r, echo=FALSE}
set.seed(123)
x = arima.sim(model = list(ar = c(1.5, -.75)), 
              n = 144)
plot(x)
```

## Is series stationary?

- Visually series appears to be stationary: $d = 0$

- The Augmented Dickey-Fuller Test confirms this 

```{r}
adf.test(x)
```


## What are candidate AR/MA orders?

```{r}
tsdisplay(x)
```

## Some candidate models to try

|             | **p** | **q** |   |   |
|-------------|-------|-------|---|---|
| **model 1** | 1     | 0     |   |   |
| **model 2** | 0     | 1     |   |   |
| **model 3** | 2     | 0     |   |   |

```{r}
m1 = Arima(x, order = c(1, 0, 0))
m2 = Arima(x, order = c(0, 0, 1))
m3 = Arima(x, order = c(2, 0, 0))
```

## ARIMA(1, 0, 0)

:::: {.columns}

::: {.column width="50%"}
```{r}
m1
```
:::

::: {.column width="50%"}
```{r}
acf(resid(m1))
```
:::

::::

## ARIMA(0, 0, 1): 

:::: {.columns}

::: {.column width="50%"}
```{r}
m2
```
:::

::: {.column width="50%"}
```{r}
acf(resid(m2))
```
:::

::::

## ARIMA(2, 0, 0)

:::: {.columns}

::: {.column width="50%"}
```{r}
m3
```
:::

::: {.column width="50%"}
```{r}
acf(resid(m3))
```
:::

::::

## Important note

- This was simulated data from an AR(2) model!
- Real data will NEVER be this simple and clean model
- Some ACF lags will be hard to interpret
- AIC/BIC values will be much closer together

# ---- Fit SARIMA(p, d, q)(P, D, Q)

## Let's try it out

Recall how to download data from the FRED:

```{r, echo=FALSE}
library(fredr)
fredr_set_key("052142bc981666b4ebcb1c8df98d006b")
```

```{r}
library(fredr)
# fredr_set_key("your_key_here")

dat = fredr(series_id = "MSACSR") |> 
  select(time = date, value) 

ts_plot(dat)
```

## Clearly need d=1, What about AR/MA/SAR/SMA orders?

:::: {.columns}

::: {.column width="50%"}
```{r}
x = ts_ts(dat)
acf(diff(x), lag.max = 12*5)
```
:::

::: {.column width="50%"}
```{r}
pacf(diff(x), lag.max = 12*5)
```
:::

::::

## Lets try SARIMA(0, 1, 1)(0, 0, 1)

```{r}
fit = Arima(x, order = c(0, 1, 1), seasonal = c(0, 0, 1))
acf(resid(fit))
```

## Lets try SARIMA(0, 1, 1)(0, 1, 1)

Might need to add seasonal difference, $D=1$

```{r}
fit2 = Arima(x, order = c(0, 1, 1), seasonal = c(0, 1, 1))
acf(resid(fit2))
```

Moderately better, maybe...

## Lets try SARIMA(0, 1, 1)(1, 1, 0)

```{r}
fit3 = Arima(x, order = c(0, 1, 1), seasonal = c(1, 1, 0))
acf(resid(fit3))
```

Looks worse!

