---
title: "2026 Time Series"
author: "07-smoothing"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
editor_options: 
  chunk_output_type: console
---

::: {.hidden}
$$
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\newcommand\mcy{{\bf y}}
$$
:::

```{r, include=FALSE}
library(forecast)
library(fpp3)
library(tidyverse)
library(tsbox)
library(zoo)
library(seasonal)
library(astsa)
library(patchwork)
library(tseries)
```

```{r setup, include=FALSE}
# Set up chunk for all slides
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  dev.args = list(pointsize = 11)
)
```


# ---- Smoothing Splines

## Basics before smoothing splines

- An obvious way to smooth data would be to fit a polynomial regression in terms of time.

$$ x_t = \beta_0 + \beta_1 t + \beta_2 t^2 + \beta_3 t^3 + w_t $$

- An extension of polynomial regression is to first divide time $t = 1, \ldots, n$
into $k$ intervals where $t_1, \ldots, t_k$ are called _knots_. 

  * Fit polynomial regression on each interval. This is called **cubic splines**

## Smoothing Splines

- Minimizes a compromise between the fit and the degree of smoothness given by
$$ \sum_{t=1}^{n}[x_t - m_t]^2 + \lambda \int (m_t^{\prime\prime})^2 dt $$
where $m_t$ is a cubic spline with knot at each $t$. $\lambda$ controlls the degree of smoothness


## Smoothing spline story time

Think of taking a long drive where $m_t$ is the position of your car at time $t$. In this case, $m_t^{''}$ is instantaneous acceleration/deceleration, and $\int (m_t^{''})^2 \, dt$ is a measure of the total amount of acceleration and deceleration on your trip. A smooth drive would be one where a constant velocity is maintained (i.e., $m_t^{''} = 0$). A choppy ride would be when the driver is constantly accelerating and decelerating, such as beginning drivers tend to do.

If $\lambda = 0$, we don’t care how choppy the ride is, and this leads to $m_t = x_t$, the data, which are not smooth. If $\lambda = \infty$, we insist on no acceleration or deceleration ($m'' = 0$); in this case, our drive must be at constant velocity, $m_t = c + vt$, and consequently very smooth. Thus, $\lambda$ is seen as a trade-off between linear regression (completely smooth) and the data itself (no smoothness). The larger the value of $\lambda$, the smoother the fit.

## Smoothing spline in R

- In R, the smoothing parameter is called `spar` and it is monotonically related to $\lambda$

- type `?smooth.spline` to view the help file for details.

```{r}
plot(soi)
lines(smooth.spline(time(soi), soi, spar=.5), lwd=2, col=4)
lines(smooth.spline(time(soi), soi, spar= 1), lty=2, lwd=2, col=2)
```

## Interpolating Missing values

```{r}
n = 30
t = 1:n
c = sin(2*pi*t / 10)
x = c + rnorm(n)
plot(x, type = "b")
lines(c, col = 'grey')
```

##

- Inject a missing value

```{r}
x[15] = NA
plot(x, type = "b")
abline(v = 15, lty = 'dotted')
```

##

- fit basic smooth spline, note need to drop NA index

```{r}
ss = smooth.spline(x = time(x[-15]), y = x[-15]) # auto select spar
plot(x, type = "b")
lines(ss, col = 2)
points(ss$x[15], ss$y[15], col = 2, pch = 19)
```

## Using `na.spline` from zoo 

- No need to find index of NA values

```{r}
plot(x, type = "b")
points(ss$x[15], ss$y[15], col = 2, pch = 19)
points(na.spline(x), col = 3, pch = 19)
```

# ---- Simple Exponential Smoothing

## Basic Idea

- Naive method 

$$ \widehat{y}_{T+h | T} = y_T $$

- Average Method

$$ \widehat{y}_{T+h | T} = \frac{1}{T} \sum_{t=1}^{T} y_t $$


These are both quite extreme forecasts!


##

- We often want something between these two extremes. 

- Attach larger weights to more recent observations than to observations from the distant past. 

- This is exactly the concept behind simple **exponential smoothing**

$$
\widehat{y}_{T+1 | T} 
= 
\alpha y_T + \alpha(1-\alpha)y_{T-1} + \alpha(1-\alpha)^2 y_{T-2} + \cdots 
$$

where $0 \leq \alpha \leq 1$

## 

- We will look at two equivalent forms of simple exponential smoothing

1. Weighted average form

1. Component form

## Weighted average form 

- The forecast at time $T+1$ is equal to a weighted average between the most recent observation $y_T$ and the previous forecast $\widehat{y}_{T|T-1}$

$$
\hat{y}_{T+1|T} = \alpha y_T + (1-\alpha) \hat{y}_{T|T-1}
$$

## Component Form

- For simple exponential smoothing, the only component included is the level $\ell_t$

$$
\begin{align*}
  \text{Forecast equation}  && \hat{y}_{t+h|t} & = \ell_{t}\\
  \text{Smoothing equation} && \ell_{t}        & = \alpha y_{t} + (1 - \alpha)\ell_{t-1},
\end{align*}
$$

## Optimisation

- For simple exponential smoothing, we need to select the values of $\alpha$ and $\ell_0$.

- the unknown parameters and the initial values for any exponential smoothing method can be estimated by minimising the SSE

$$
\text{SSE}=\sum_{t=1}^T(y_t - \hat{y}_{t|t-1})^2=\sum_{t=1}^Te_t^2.
$$

# ---- Adding Trend and Seasonality 

## Holt’s linear trend method {.smaller}

- This method involves a forecast equation and two smoothing equations (one for the level and one for the trend)

$$
\begin{align*}
  \text{Forecast equation}&& \hat{y}_{t+h|t} &= \ell_{t} + hb_{t} \\
  \text{Level equation}   && \ell_{t} &= \alpha y_{t} + (1 - \alpha)(\ell_{t-1} + b_{t-1})\\
  \text{Trend equation}   && b_{t}    &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 -\beta^*)b_{t-1},
\end{align*}
$$

- $\ell_t$  denotes an estimate of the level of the series at time t

- $b_t$ denotes an estimate of the trend (slope) of the series at time t.
 
- $\alpha$ is the smoothing parameter

- $\beta^*$ smoothing parameter for the trend

## Damped trend methods

- This method also includes a damping parameter $0 < \phi < 1$

- If $\phi = 1$ the method is identical to Holt’s linear method

- $\phi$ dampens the trend so that it approaches a constant some time in the future

$$
\begin{align*}
  \hat{y}_{t+h|t} &= \ell_{t} + (\phi+\phi^2 + \dots + \phi^{h})b_{t} \\
  \ell_{t} &= \alpha y_{t} + (1 - \alpha)(\ell_{t-1} + \phi b_{t-1})\\
  b_{t} &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 -\beta^*)\phi b_{t-1}.
\end{align*}
$$

## Example: Australian population

```{r, echo=FALSE}
aus_economy <- global_economy |>
  filter(Code == "AUS") |>
  mutate(Pop = Population / 1e6)
autoplot(aus_economy, Pop) +
  labs(y = "Millions", title = "Australian population")
```

##

- code to fit both Holt's linear trend method and the dampened trend model

```{r}
#| output-location: slide
aus_economy |>
  model(
    `Holt's method` = ETS(Pop ~ error("A") +
                       trend("A") + season("N")),
    `Damped Holt's method` = ETS(Pop ~ error("A") +
                       trend("Ad", phi = 0.9) + season("N"))
  ) |>
  forecast(h = 15) |>
  autoplot(aus_economy, level = NULL) +
  labs(title = "Australian population",
       y = "Millions") +
  guides(colour = guide_legend(title = "Forecast"))
```

## Adding Seasonality 

- Holt (1957) and Winters (1960) extended Holt’s method to capture seasonality

- The Holt-Winters seasonal method comprises the forecast equation and three smoothing equations 
  * one for the level $\ell_t$ (smoothing param $\alpha$)
  * one for the trend $b_t$ (smoothing param $\beta^*$)
  * one for the seasonal component $s_t$ (smoothing param $\gamma$)
  * let $m$ to denote the period of the seasonality
  

## Additive Holt-Winters Model
  
$$
\begin{align*}
  \hat{y}_{t+h|t} &= \ell_{t} + hb_{t} + s_{t+h-m(k+1)} \\
  \ell_{t} &= \alpha(y_{t} - s_{t-m}) + (1 - \alpha)(\ell_{t-1} + b_{t-1})\\
  b_{t} &= \beta^*(\ell_{t} - \ell_{t-1}) + (1 - \beta^*)b_{t-1}\\
  s_{t} &= \gamma (y_{t}-\ell_{t-1}-b_{t-1}) + (1-\gamma)s_{t-m},
\end{align*}
$$

## Multiplicative Holt-Winters Model

$$
\begin{align*}
  \hat{y}_{t+h|t} &= (\ell_{t} + hb_{t})s_{t+h-m(k+1)} \\
  \ell_{t} &= \alpha \frac{y_{t}}{s_{t-m}} + (1 - \alpha)(\ell_{t-1} + b_{t-1})\\
  b_{t} &= \beta^*(\ell_{t}-\ell_{t-1}) + (1 - \beta^*)b_{t-1}                \\
  s_{t} &= \gamma \frac{y_{t}}{(\ell_{t-1} + b_{t-1})} + (1 - \gamma)s_{t-m}.
\end{align*}
$$

## Example: Domestic overnight trips in Australia

```{r}
#| output-location: slide
aus_holidays <- tourism |>
  filter(Purpose == "Holiday") |>
  summarise(Trips = sum(Trips)/1e3)
fit <- aus_holidays |>
  model(
    additive = ETS(Trips ~ error("A") + trend("A") +
                                                season("A")),
    multiplicative = ETS(Trips ~ error("M") + trend("A") +
                                                season("M"))
  )
fc <- fit |> forecast(h = "3 years")
fc |>
  autoplot(aus_holidays, level = NULL) +
  labs(title="Australian domestic tourism",
       y="Overnight trips (millions)") +
  guides(colour = guide_legend(title = "Forecast"))
```

# ---- Forecasting

## Forecasting

- Point forecasts can be obtained from the models by iterating the equations 

- For example, $\widehat{y}_{T+2|T} = \ell_T + 2b_T$ and so on.

- Using `forecast()` function from **fable** package:

```{r}
#| output-location: slide
fit |>
  forecast(h = 8) |>
  autoplot(aus_holidays)+
  labs(title="Australian domestic tourism",
       y="Overnight trips (millions)")
```

## Prediction Intervals

- The prediction intervals will differ between models with additive and multiplicative methods

- For most ETS models, a prediction interval can be written as:

$$
\widehat{y}_{T+h | T} \pm c\sigma_h
$$

where $c$ depends on coverage probability and $\sigma_h$ is the forecast variance. 

- See Table 9.8 in fpp3 for a list of $\sigma_h$ values for different ETS models.

# ---- Smoothing vs. ARIMA

## Two different philosophies

- **Exponential smoothing (ETS)**: describe the data with a small set of
  interpretable components -- level, trend, season -- each updated by its own
  smoothing equation. The model is chosen by *picking a structure* (error/trend/season
  type) that matches what you see in the data.

- **ARIMA / regARIMA**: describe the *autocorrelation* of a (differenced) series
  directly, via $p, d, q$ (and seasonal $P, D, Q$). The model is chosen by
  *identifying orders* from ACF/PACF (or a numerical search) after making the series
  stationary.

- Both are legitimate ways to answer the same question -- "what's a good forecast for
  $y_{T+h}$?" -- but they get there very differently, and that difference shows up in
  when each one shines.

## When smoothing methods (ETS) tend to win

- **Many series to forecast quickly.** ETS has few, well-understood structural choices
  (error/trend/season $\times$ additive/multiplicative/damped) and fits fast --
  ideal for forecasting thousands of SKUs, store locations, or sensors at once.

- **Short series.** Fewer parameters to estimate than a comparable SARIMA model, so ETS
  can still produce sensible forecasts when there isn't enough history to reliably
  identify ARMA orders.

- **The series is dominated by level/trend/season and not much else.** If, once you
  account for trend and season, what's left really does look like noise, there's
  nothing for an ARMA error structure to add.

- **You want a forecast, not necessarily an explanation.** ETS components (level,
  trend, seasonal) are easy to describe to a non-technical audience; ARIMA
  coefficients are not.

- **You don't have (or don't want to build) covariates.** Both families can be
  univariate, but ETS is *designed* to be used that way, whereas the regARIMA approach
  from the previous lectures assumes you're supplying meaningful regressors.

## When ARIMA-based methods tend to win

- **The data show genuine, structured autocorrelation beyond level/trend/season.**
  ARIMA's whole purpose is modeling that dependence explicitly (recall the AR/MA
  ACF-PACF signatures from 05a) -- ETS has no mechanism for it.

- **You need to incorporate covariates/regressors.** regARIMA (06-regARIMA) lets you
  combine a mean-structure regression with an explicitly modeled ARMA error process;
  ETS does not have a clean regression-with-correlated-errors analogue.

- **You want principled, likelihood-based model comparison.** AICc-based comparison of
  candidate $(p,d,q)(P,D,Q)_m$ specifications, plus Ljung-Box diagnostics, gives a
  rigorous framework for defending *why* one model beats another (05a's whole
  workflow). ETS model selection is comparatively coarser (choose among $\sim$30
  error/trend/season combinations).

- **Long, information-rich series where you can afford careful modeling.** Identifying
  ARMA orders, checking residuals, and refitting is more labor-intensive per series --
  worth it when getting one series right really matters.

- **You need a fully specified stochastic (not just point-forecast) model** -- e.g.
  for simulation, spectral analysis, or downstream statistical inference that depends
  on the assumed data-generating process.

## Pros and cons at a glance

| | **Smoothing (ETS)** | **ARIMA / regARIMA** |
|---|---|---|
| **Interpretability** | High -- level/trend/season components | Lower -- AR/MA coefficients are less intuitive |
| **Speed / automation at scale** | Fast, few structural choices, easy to batch | Slower -- identification + diagnostics per series |
| **Handles non-integer/complex seasonality** | Limited (needs integer $m$; struggles with multiple seasonalities without extensions) | Same integer-$m$ limitation, but Fourier terms in the regression part sidestep it (05a/06) |
| **Covariates / exogenous regressors** | Not native to the framework | Natural via regARIMA (`xreg`, `ARIMA(y ~ x + pdq())`) |
| **Captures short-term serial correlation beyond trend/season** | No explicit mechanism | Yes -- that's the point of the ARMA error |
| **Data requirements** | Works reasonably well on short series | Needs more data to reliably identify orders, especially seasonal ones |
| **Model selection rigor** | Coarse (choose among ETS structures via AICc) | Fine-grained (search over $p,q,P,Q$ via AICc + residual diagnostics) |
| **Risk of overfitting** | Lower (few parameters) | Higher with unconstrained grid/`auto.arima` searches -- always check residuals, not just AICc |

## Practical guidance

1. **Plot the data first**, regardless of which family you're leaning toward -- trend,
   seasonality, and obvious anomalies should drive the choice, not habit.
2. If you mainly need **trend + seasonal extrapolation** for **many series, fast** --
   start with **ETS**.
3. If residuals from a simple mean-structure regression (or from an ETS fit) still show
   **clear autocorrelation**, or you have **covariates** you want to use properly --
   move to **ARIMA / regARIMA**.
4. In practice, **fit both and compare** out-of-sample (e.g. rolling-origin CV from
   05a) -- `fable` makes this nearly free, since `ETS()` and `ARIMA()` share the same
   `model()`/`forecast()`/`accuracy()` interface.
5. Remember: **AICc is not comparable between ETS and ARIMA models fit to different
   representations of the data** (e.g. levels vs. differences) -- always validate the
   final choice with residual diagnostics and/or cross-validated forecast accuracy, not
   information criteria alone.





