---
title: "2026 Time Series"
author: "06-regARIMA"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
editor_options: 
  chunk_output_type: console
---

::: {.hidden}
$$
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\newcommand\mcy{{\bf y}}
$$
:::

```{r, include=FALSE}
library(forecast)
library(fpp3)
library(tidyverse)
library(tsbox)
library(zoo)
library(seasonal)
library(astsa)
library(patchwork)
library(tseries)
```

```{r setup, include=FALSE}
# Set up chunk for all slides
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  dev.args = list(pointsize = 11)
)
```

# RegARIMA

## regARIMA

- We have discussed the <ins>classical regression</ins> model with uncorrelated errors

$$ y_t = \beta_0 + \beta_1 x_{1, t} + \cdots + \beta_r x_{r, t} + \epsilon_t $$

- Consider not the regression model where the error term, $x_t$ has some ACVF
function $\gamma(h)$

$$ y_t = \sum_{j=1}^{r}\beta_j z_{t, j} + x_t $$

## Matrix form and estimator

$$ \textbf{y} = \textbf{Z}\boldsymbol{\beta} + \textbf{x} $$

We have derived 

$$ \widehat{\beta} = 
(Z^\prime \Gamma^{-1} Z)^{-1} Z^\prime \Gamma^{-1} y $$

## ARMA Errors

- The model needs the an ACVF function, and resulting $\Gamma$ to be fit.

- It is often possible to assume a stationary covariance structure for the error 
process that corresponds to a linear process and try to find an ARMA 
representation.

- This amounts to deciding on a $p$ and $q$, but how do we decide?

## Example

```{r}
par(mfrow=c(3,1)) # plot the data
plot(cmort, main="Cardiovascular Mortality", xlab="", ylab="") 
plot(tempr, main="Temperature", xlab="", ylab="")
plot(part, main="Particulates", xlab="", ylab="")
```

## Example

```{r}
pairs(
  cbind(
    Mortality=cmort, 
    Temperature=tempr, 
    Particulates=part
  )
) 
```

## Example

- Consider model with <ins>uncorrelated</ins> errors

$$ M_t = 
\beta_1 + \beta_2 t + \beta_3 T_t + \beta_4 T_t^2 + \beta_5 P_t + \epsilon_t $$

```{r}
temp = tempr-mean(tempr) # mean center temp
temp2 = temp^2
trend = time(cmort) # time
fit = lm(cmort~ trend + temp + temp2 + part, na.action=NULL) 
summary(fit) # regression results
```

## AR and MA behavior

Before we look at the residuals of the model and decide on what type of ACVF
behavior we see. Let's look at known behavior:

|       | AR(p)                 | MA(q)                     | 
|-------|-----------------------|---------------------------|
| ACF   | Tails off             | PACF Cuts off after lag p |
| PACF  | Cuts off after lag p  | Tails off                 |


## Example

```{r}
tsdisplay(residuals(fit), lag.max = 30)
```

## Example

- Would have been a mistake to fit an error with uncorrelated residuals

- A plot of the residuals indicates an AR(2) stochastic structure

$$ M_t = 
\beta_1 + \beta_2 t + \beta_3 T_t + \beta_4 T_t^2 + \beta_5 P_t + x_t $$ 

where 

$$ x_t = \phi_1 x_{t-1} + \phi_2 x_{t-2} + w_t $$

## Example 

- There are many R functions that will fit an ARMA model and allow regressors

```{r}
regarima_fit <- sarima(cmort, 2,0,0, xreg=cbind(trend,temp,temp2,part), details = FALSE)
```

## Example: residuals of regARIMA fit

```{r}
tsdisplay(resid(regarima_fit$fit), lag.max = 30)
```

## Overall proceedure

1. Run OLS acting as if the errors are uncorrelated
1. Identify ARMA model(s) for the residuals 
1. Run WLS (or MLE) on the regression model with ACVF of specified model
1. Inspect residuals of regARIMA model for whiteness, adjust if needed

## Automatic modeling 

- fpp3 text has some built in helper function for fitting regARIMA

- To fit the regARIMA model we just fit, with AR(2) error we could use

```{r, eval=FALSE}
ARIMA(y ~ trend + temp + temp2 + part + pdq(2,0,0))
```

- If no `pdq()` is specified, automatic model choice will be made.


## US personal consumption and Income

- We would like to forecast changes in expenditure based on changes in income.

```{r, echo=FALSE}
us_change |>
  pivot_longer(c(Consumption, Income),
               names_to = "var", values_to = "value") |>
  ggplot(aes(x = Quarter, y = value)) +
  geom_line() +
  facet_grid(vars(var), scales = "free_y") +
  labs(title = "US consumption and personal income",
       y = "Quarterly % change")
```

## Model fit 

```{r}
fit <- us_change |>
  model(ARIMA(Consumption ~ Income))
report(fit)
```

## Model fit residuals (regARIMA and OLS)

```{r, echo=FALSE}
bind_rows(
    `Regression residuals` =
        as_tibble(residuals(fit, type = "regression")),
    `ARIMA residuals` =
        as_tibble(residuals(fit, type = "innovation")),
    .id = "type"
  ) |>
  mutate(
    type = factor(type, levels=c(
      "Regression residuals", "ARIMA residuals"))
  ) |>
  ggplot(aes(x = Quarter, y = .resid)) +
  geom_line() +
  facet_grid(vars(type))
```

# Forecasting 

## Electricity Demand

- Daily electricity demand can be modelled as a function of temperature

- more electricity is used on cold days due to heating and hot days due to air conditioning

- Moderate temperature requires the least amount of electricity
    * This leads to both linear and non-linear dynamics


## Electricity Demand

```{r, echo=FALSE}
vic_elec_daily <- vic_elec |>
  filter(year(Time) == 2014) |>
  index_by(Date = date(Time)) |>
  summarise(
    Demand = sum(Demand) / 1e3,
    Temperature = max(Temperature),
    Holiday = any(Holiday)
  ) |>
  mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"
  ))

vic_elec_daily |>
  ggplot(aes(x = Temperature, y = Demand, colour = Day_Type)) +
  geom_point() +
  labs(y = "Electricity demand (GW)",
       x = "Maximum daily temperature")
```

## Electricity Demand

```{r}
vic_elec_daily |>
  pivot_longer(c(Demand, Temperature)) |>
  ggplot(aes(x = Date, y = value)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") + ylab("")
```


## Electricity Demand

```{r}
fit <- vic_elec_daily |>
  model(ARIMA(Demand ~ Temperature + I(Temperature^2) +
                (Day_Type == "Weekday")))
report(fit)
```

## Electricity Demand

```{r}
fit |> gg_tsresiduals()
```


## Electricity Demand

Conduct a LBQ test for significant autocorrelation in the residuals 

```{r}
Box.test(resid(fit)$.resid, lag = 14)
```

## Electricity Demand

- There is clear heteroscedasticity in the residuals, with higher variance in January and February, and lower variance in May.
- The model also has some significant autocorrelation in the residuals, and the histogram of the residuals shows long tails.
- All of these issues with the residuals may affect the coverage of the prediction intervals, but the point forecasts should still be ok.

## Electricity Demand: Forecast

- Using the estimated model we **forecast 14 days ahead**
- Starting from Thursday 1 January 2015 (a non-work-day being a public holiday for New Years Day).
- In this case, we could obtain weather forecasts from the weather bureau for the next 14 days. 
    * But for the sake of illustration, we will use scenario based forecasting (as introduced in Section 7.6 of fpp3) where we set the temperature for the next 14 days to a constant 26 degrees.
- Working day dummy variables are set to known future values.

## Electricity Demand: Forecast

```{r}
#| output-location: slide
vic_elec_future <- new_data(vic_elec_daily, 14) |>
  mutate(
    Temperature = 26,
    Holiday = c(TRUE, rep(FALSE, 13)),
    Day_Type = case_when(
      Holiday ~ "Holiday",
      wday(Date) %in% 2:6 ~ "Weekday",
      TRUE ~ "Weekend"
    )
  )
forecast(fit, vic_elec_future) |>
  autoplot(vic_elec_daily) +
  labs(title="Daily electricity demand: Victoria",
       y="GW")
```

