---
title: "Midterm: Slide Deck"
author: "Emily Fagerstrom"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
library(tuneR)
library(tidyverse)
library(ggplot2)
library(gganimate) #for the animation with the sound
library(forecast)
library(gt) # chat used it to make the table of notes/freqs look nice



```


# Cross Correlation

## Stereo Sound

- Two independent time series ($L_t$ and $R_t$) trick your brain 
- Share a lot of information (usually all instruments sent to both)
- Thus, waveforms look close to identical; minor tweaks

## Monocompatibility

- If left and right channels get out of phase, the waves can cancel each other
- If song is played on a smartphone or bluetooth speaker (in mono), sound can vanish
- Correlation analysis to check mixes are "mono-compatible"

## Live vs Recorded vs Remastered

- Download Goo Goo Dolls "Slide" from youtube
- Original
- "Remastered"
- Live in Buffalo July 4, 2004


![](GooGooDollsIris-1){width=50%}

## Fun fact!  44100 Hz

- Limit for healthy hearing: $\approx$ 20 kHz
- Nyquist-Shannon Sampling Them: sample at more than twice the highest frequency you want to record
- Sound engineers want a little bit of "breathing" room (especially with 1970s analog technology)
- American and European TV were different (30 frames per second vs 25)
- Want a number with a lot of factors to cleanly divide audio samples across video frames
- $44,100 = 2^2 \times 3^2 \times 5^2 \times 7^2$
- Sony pitched Compact Disc standard in 1980; locked in now

## CCFs (left and right channels)

```{r}
#| echo: false

song_og <- readMP3("Slide.mp3")
song_remastered <- readMP3("Slide-Remastered.mp3")
song_live <- readMP3("Slide-Live.mp3")

# Sampling rate (double check it is the usual 44100 Hz)
fs <- song_og@samp.rate
print(paste("Sampling Rate:", fs, "Hz"))

# Tiny smidgin of chorus: "yeaaaah we're gonna let it slide"
start_sec <- 44.2
duration_sec <- 3

# Sample indices
start_sample <- start_sec * fs
end_sample <- (start_sec + duration_sec) * fs

# Left and Right channels
og_left <- song_og@left[start_sample:end_sample]
og_right <- song_og@right[start_sample:end_sample]

#L/R Live version
live_left <- song_live@left[start_sample:end_sample]
live_right <- song_live@right[start_sample:end_sample]

#L/R Remastered version
rem_left <- song_remastered@left[start_sample:end_sample]
rem_right <- song_remastered@right[start_sample:end_sample]




chorus_stereo <- Wave(left = og_left, 
                      right = og_right, 
                      bit = 16, 
                      samp.rate = fs)

writeWave(chorus_stereo, filename = "yeeaaah.wav")


```


<audio controls>
  <source src="yeeaaah.wav" type="audio/wav">
</audio>


```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 9
#| out-width: "100%"
#| echo: false

par(mfrow = c(3, 1))

ccf_og <- ccf(og_left, og_right, lag.max = 100, 
               main = "CCF: Original",
               ylab = "CCF", col = "blue", lwd = 2)

ccf_rem <- ccf(rem_left, rem_right, lag.max = 100, 
               main = "CCF: Remastered Version",
               ylab = "CCF", col = "red", lwd = 2)

ccf_live <- ccf(live_left, live_right, lag.max = 100, 
                main = "CCF: Live Version",
                ylab = "CCF", col = "green", lwd = 2)

# reset layout
par(mfrow = c(1, 1))

```


## CCFs: Remastered

![](Loudness-wiki){width=75%}


# Autocorrelation

## Note Detection using Rolling Variance

- Intro (first 9 seconds)
- Look at *small* snippets (1024 frames; .23 seconds each)
- Check the variance of each piece
- Places with high variance are likely to be the onset of the next note

## Note Detection using Rolling Variance

```{r}
#| echo: false

rm(list = setdiff(ls(), c("song_og","fs")))

# audio clip (left channel; intro)
intro_left <- song_og@left[1:(44100 * 9)]
```

```{r}
#| eval: false
#| echo: false

# Run this code to create the .mov, but then don't actually run it during the slides, just grab the animation it produced

writeWave(Wave(left = intro_left, samp.rate = 44100, bit = 16), "intro_clip.wav")

# rolling variance dataframe
# (doesn't need to be power of 2 for this purpose;
# but that is usually how things are chunked in audio world to allow for FFTs) 
frame_size <- 1024
n_frames <- floor(length(intro_left) / frame_size)
frames <- matrix(intro_left[1:(n_frames * frame_size)], nrow = frame_size)
frame_vars <- apply(frames, 2, var)
frame_times <- (seq_along(frame_vars) * frame_size) / 44100
df_variance <- data.frame(time = frame_times, variance = frame_vars)

# rolling variance plot and animation
p <- ggplot(df_variance, aes(x = time, y = variance)) +
  geom_line(color = "purple", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Note Onset Detection: Intro Riff",
    subtitle = "Time: {frame_along} seconds",
    x = "Time (Seconds)", 
    y = "Variance (Signal Energy)"
  )

anim <- p + transition_reveal(time)

# combine animation and audio
animate(anim, 
        nframes = 180, 
        fps = 20, 
        renderer = av_renderer(file = "variance_video.mp4", audio = "intro_clip.wav"))

```


{{< video variance_video.mp4 width="700" >}}


## Quick digression: music frequencies

- Pure sine wave
- $y(t) = \sin(t)$: period $2\pi$, frequency $1/2\pi$
- $y(t) =\sin(2 \pi t)$: period $1$, frequency $1$
- $y(t) = \sin( 2 \pi \cdot 440 t)$: frequency 440, "440 A"

A 440 Hz sound wave (sine)

<audio controls>
  <source src="440Hz.wav" type="audio/wav">
</audio>


## Quick digression: musical frequencies

- Double the frequency
- Same note ("pitch") but higher

A 220 Hz sound wave (sine)

<audio controls>
  <source src="220Hz.wav" type="audio/wav">
</audio>

A 440 Hz sound wave (sine)

<audio controls>
  <source src="440Hz.wav" type="audio/wav">
</audio>



```{r}
#| fig-width: 8
#| fig-height: 4.5
#| echo: false

library(ggplot2)

t <- seq(0, 0.005, by = 0.00001)

waves <- data.frame(
  time = rep(t, 2),
  amplitude = c(
    sin(2 * pi * 220 * t),
    sin(2 * pi * 440 * t)
  ),
  frequency = factor(
    rep(c(220, 440), each = length(t)),
    levels = c(220, 440),
    labels = c("220 Hz", "440 Hz")
  )
)

ggplot(waves, aes(x = time, y = amplitude, color = frequency)) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    values = c("220 Hz" = "blue", "440 Hz" = "red")
  ) +
  labs(
    title = "Pure Sine Waves",
    subtitle = "Comparing 220 Hz and 440 Hz",
    x = "Time (seconds)",
    y = "Amplitude",
    color = "Frequency"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 17),
    plot.subtitle = element_text(color = "gray40"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```


![](piano){width=75%}


## Quick (I swear) digression: musical frequencies

- Rest of the notes
- 12 notes that repeat
![](numbered-notes){width=75%}


$$\mathrm{frequency \,\, n \,\, notes \,\, above \,\, or \,\, below \,\, A: \,\,\,\,}  440 \cdot 2^{n/12} $$

```{r}
#| echo: false


n <- seq(1,24,1)

n <- 1:24

notes <- rep(
  c("B♭", "B", "C", "D♭", "D", "E♭", 
    "E", "F", "G♭", "G", "A♭", "A"),
  2
)

freqs <- round(110 * 2^(n / 12), 3)

df <- data.frame(
  Note = notes,
  Frequency = freqs
)

slide_table <- data.frame(
  Note_1 = df$Note[1:12],
  Hz_1   = df$Frequency[1:12],
  Note_2 = df$Note[13:24],
  Hz_2   = df$Frequency[13:24]
)

slide_table |>
  gt() |>
  cols_label(
    Note_1 = "Note",
    Hz_1   = "Frequency (Hz)",
    Note_2 = "Note",
    Hz_2   = "Frequency (Hz)"
  ) |>
  tab_header(
    title = "Frequencies of Notes on a Piano"
  ) |>
  fmt_number(
    columns = c(Hz_1, Hz_2),
    decimals = 3
  ) |>
  tab_options(
    table.font.size = 18,
    heading.title.font.size = 24,
    column_labels.font.weight = "bold",
    data_row.padding = px(8)
  )
```



## Pitch Detection using ACF

First note of the song (.7 to .9 sec)

```{r}
# Starting note
note_start_sec <- .7
note_end_sec <- .9

note_segment <- intro_left[(note_start_sec * 44100):(note_end_sec * 44100)]
writeWave(Wave(left = note_segment, samp.rate = 44100, bit = 16), "first_note.wav")

# ACF
acf(note_segment, lag.max = 300, 
    main = "ACF (a single note)",
    col = "blue", lwd = 2)

# PACF
pacf(note_segment, lag.max = 300, 
    main = "PACF (a single note)",
    col = "blue", lwd = 2)

```




```{r}
#| eval: false
#| echo: false


freq <- 175.7
duration <- 1
sample_rate <- 44100

t <- seq(0, duration, 1 / sample_rate)
wave <- sin(2 * pi * freq * t)

wav <- Wave(
  left = as.integer(wave * 32767), #this is max volume in 16-bit audio
  samp.rate = sample_rate,
  bit = 16
)

writeWave(wav, "176Hz.wav")


freq <- 440
duration <- 1
sample_rate <- 44100

t <- seq(0, duration, 1 / sample_rate)
wave <- sin(2 * pi * freq * t)

wav <- Wave(
  left = as.integer(wave * 32767), 
  samp.rate = sample_rate,
  bit = 16
)

writeWave(wav, "440Hz.wav")


freq <- 220
duration <- 1
sample_rate <- 44100

t <- seq(0, duration, 1 / sample_rate)
wave <- sin(2 * pi * freq * t)

wav <- Wave(
  left = as.integer(wave * 32767), 
  samp.rate = sample_rate,
  bit = 16
)

writeWave(wav, "220Hz.wav")

```


## Pitch Detection using ACF
- At lag 251, autocorrelation is almost 1
- $\frac{251 \,\, \mathrm{cycles}}{1} \times \frac{1 \,\,\mathrm{second}}{44100 \,\,\mathrm{cycles}} = .00569 \,\, \mathrm{sec}$
- Frequency is 44100/251
- $44100/251 \approx 175.697 \approx F$


## Pitch Detection using ACF

The intro (left channel)

<audio controls>
  <source src="intro_clip.wav" type="audio/wav">
</audio>

The first note (.7 to .9 sec)

<audio controls>
  <source src="first_note.wav" type="audio/wav">
</audio>

A 175.7 Hz sound wave (sine)

<audio controls>
  <source src="176Hz.wav" type="audio/wav">
</audio>


## Idea: 

- Use rolling variance to determine note onset
- Find local maxima; consider neighborhood to the right of the max
- Within that neighborhood, run ACF
- Find local max of the ACF, $m$
- The note has freq $44100/m$


## Idea: Underwhelming

```{r}
#| echo: false

rm(list = setdiff(ls(), c("song_og","fs")))

intro_left <- song_og@left[1:(44100 * 9)]

frame_size <- 1024
n_frames <- floor(length(intro_left) / frame_size)
frames <- matrix(intro_left[1:(n_frames * frame_size)], nrow = frame_size)
frame_vars <- apply(frames, 2, var)
frame_times <- (seq_along(frame_vars) * frame_size) / fs


# Function to find local peaks in variance (Note Onsets)
find_peaks <- function(x, threshold_quantile = 0.85, span = 10) {
  thresh <- quantile(x, threshold_quantile, na.rm = TRUE)
  peaks <- c()
  for (i in (span + 1):(length(x) - span)) {
    if (x[i] == max(x[(i - span):(i + span)], na.rm = TRUE) && x[i] > thresh) {
      peaks <- c(peaks, i)
    }
  }
  return(peaks)
}

onset_indices <- find_peaks(frame_vars, threshold_quantile = 0.82, span = 15)
onset_times <- frame_times[onset_indices]

# Find the FIRST local maximum in the ACF
find_first_local_max <- function(vals, min_threshold = 0.25) {
  for (i in 2:(length(vals) - 1)) {
    # Check if it's a peak (higher than neighbors) and crosses a baseline correlation threshold
    if (vals[i] > vals[i - 1] && vals[i] >= vals[i + 1] && vals[i] > min_threshold) {
      return(i)
    }
  }
  # Fallback to global max if no local peak meets criteria
  return(which.max(vals))
}


# Loop through each detected onset, grab the post-attack window, and run ACF
results_list <- list()

for (i in seq_along(onset_times)) {
  t_onset <- onset_times[i]
  
  # Take a 0.1-second window slightly to the right of the onset
  seg_start <- round((t_onset + 0.05) * fs)
  seg_end <- round((t_onset + 0.15) * fs)
  
  # Safety bounds check
  if (seg_end > length(intro_left) || seg_start < 1) next
  
  note_segment <- as.numeric(intro_left[seg_start:seg_end])
  
  # Use stats::acf explicitly to avoid any function/variable name collisions
  acf_res <- stats::acf(note_segment, lag.max = 800, plot = FALSE)
  lags_vec <- as.vector(acf_res$lag)
  acf_vec <- as.vector(acf_res$acf)
  
  # Restrict search to valid guitar lag range (~60 Hz to ~1000 Hz)
  search_idx <- which(lags_vec >= 40 & lags_vec <= 750)
  if (length(search_idx) == 0) next
  
  sub_lags <- lags_vec[search_idx]
  sub_acf <- acf_vec[search_idx]
  
  # Find the lag with the highest autocorrelation
  best_lag <- sub_lags[which.max(sub_acf)] # not quite; want first local max
  estimated_freq <- fs / best_lag
  
  results_list[[i]] <- data.frame(
    Note_Index = i,
    Onset_Time = round(t_onset, 2),
    Best_Lag = best_lag,
    Frequency_Hz = round(estimated_freq, 2)
  )
}

# Combine into a clean summary table
detected_notes_df <- bind_rows(results_list)
print(detected_notes_df)
```


Did pick out a lot of the A flats and E flats (though in the wrong octave), but, on further reflection, rolling variance isn't catching the high note.


{{< video variance_video.mp4 width="700" >}}



## PACF

```{r}
# Starting note, but already "ringing"
note_start_sec <- .85
note_end_sec <- .9

note_segment <- intro_left[(note_start_sec * 44100):(note_end_sec * 44100)]
writeWave(Wave(left = note_segment, samp.rate = 44100, bit = 16), "first_note.wav")

# PACF
pacf(note_segment, lag.max = 100, 
    main = "PACF (a single note, already playing)",
    col = "blue", lwd = 2)

```

## Synthesize sound via AR coefficients

Method 1

```{r}

# don't use mle - high freq problems
ar_model <- ar(note_segment, method = "burg")

print(paste("Optimal AR Order chosen by AIC:", ar_model$order))
print(ar_model)


phi <- ar_model$ar
sigma2 <- ar_model$var.pred
set.seed(42)
sustained_sim <- simulate(ar_model, nsim = round(3 * 44100)) #lasts 3 sec

# scale safely to 16-bit audio range; HAD TO CRANK THIS WAY DOWN
sustained_scaled <- as.integer(scale(as.numeric(sustained_sim)) * 4000)

writeWave(Wave(left = sustained_scaled, samp.rate = fs, bit = 16), "synth2.wav")

# Plot the spectral density implied by your AR(33) model
spectrum(note_segment, method = "ar", 
         main = "AR(33) Spectral Density (Guitar Timbre Fingerprint)",
         xlab = "Frequency (Normalized)", ylab = "Power Spectral Density")

```


COMPLETE FAILURE!!!


<audio controls>
  <source src="synth2.wav" type="audio/wav">
</audio>





```{r}

a <- ar_model$ar

# Use the last 33 samples of the real guitar as the initial state
history <- as.numeric(tail(note_segment, 33))

n_synth <- 3 * 44100

synth <- numeric(n_synth)

# Put the initial 33 samples into the beginning
synth[1:33] <- history

# Now let the AR model ring with ZERO new excitation
for (n in 34:n_synth) {
    synth[n] <- sum(a * synth[(n-1):(n-33)])
}

# Normalize without changing the waveform shape
synth <- synth / max(abs(synth))

# Leave some headroom
synth <- synth * 0.9 * 32767

synth_int <- as.integer(synth)

writeWave(
    Wave(left = synth_int, samp.rate = 44100, bit = 16),
    "synth3.wav"
)



```





# More complex sounds

## Pure sine waves vs instruments

Basic pitch plus overtones:

(synthetic example)

\begin{align}
y(t) = & 1 \sin( 2 \pi \cdot 220 t) + \\
      &  .1 \sin ( 2 \pi \cdot 330 t) +\\
      &  .8 \sin ( 2 \pi \cdot 660 t) +\\
      & .4 \sin (2 \pi \cdot 880 t)
\end{align}

Different overtones can have different phases or decay rates:

\begin{align}
y(t) = & 1 \sin[ 2 \pi \cdot 220 t] + \\
      &  .1 \sin[ 2 \pi \cdot 330 (t+.0003)] * 1.3^{-t} +\\
      &  .8 \sin[ 2 \pi \cdot 660 t]  * 1.1^{-t} +\\
      & .4 \sin[ \pi \cdot 880 (t-.0001)] * 2^{-t}
\end{align}


```{r}
# synthetic sounds
duration <- 5
sample_rate <- 44100

t <- seq(0, duration, 1 / sample_rate)

wave1 <-  1 * sin(2 * pi * 220 * t) +
         .1 * sin(2 * pi * 330 * (t+.0003)) * 1.3^(-t) +
         .8 * sin(2 * pi * 660 * t) * 1.1^(-t) +
         .4 * sin(2 * pi * 880 * (t-.0001)) * 2^(-t)

wav <- Wave(
  left = as.integer(wave1 * 15000), #just in case waves "aligned", dont want to use 32767 
  samp.rate = sample_rate,
  bit = 16
)

writeWave(wav, "synth1.wav")

```


Synthetic sound

<audio controls>
  <source src="synth1.wav" type="audio/wav">
</audio>




# Other Directions

## Other time series

- One channel of a three-minute song, is a long time series with

$3 \times 60 \times 44100 = 7,938,000$

- Instead, break into pieces of size 8192
(powers of 2 are helpful for FT)
- Each piece is about .186 seconds of song
- Aggregate within each .186 second piece

## Other time series

```{r}

# put into mono
mono_og <- rowMeans(cbind(song_og@left, song_og@right))

# get rid of silence at the beginning
first_nonzero <- which(mono_og != 0)[1]
first_chunk_start <- first_nonzero
mono_og <- mono_og[first_chunk_start:length(mono_og)]


# 2048 -> about .046 sec; 4096 -> .092; could bump to 8192?
chunk_size <- 8192
n_chunks <- floor(length(mono_og) / chunk_size)

mono_trim <- mono_og[1:(n_chunks * chunk_size)]

chunks <- matrix(
  mono_trim,
  nrow = chunk_size,
  ncol = n_chunks
)

dim(chunks)


# RMS (root mean squared) amplitude, (sqrt of sum of squares, measures overall energy)

rms <- apply(chunks, 2, function(z) {
  sqrt(mean(z^2))
})

# max amplitude

max_amp <- apply(chunks, 2, function(z) {
  max(abs(z))
})

# standard deviation / variance

sd_amp <- apply(chunks, 2, sd)
var_amp <- apply(chunks, 2, var)

# spectral centroid

spectral_centroid <- function(z, fs) {
  
  n <- length(z)
  
  # FFT
  fft_z <- fft(z)
  
  # Only keep positive frequencies
  half <- floor(n / 2) + 1
  
  magnitude <- Mod(fft_z[1:half])
  
  frequencies <- (0:(half - 1)) * fs / n
  
  sum(frequencies * magnitude) / sum(magnitude)
}

centroid <- apply(
  chunks,
  2,
  spectral_centroid,
  fs = fs
)

# spectral bandwidth

spectral_bandwidth <- function(z, fs) {
  
  n <- length(z)
  
  fft_z <- fft(z)
  
  half <- floor(n / 2) + 1
  
  magnitude <- Mod(fft_z[1:half])
  
  frequencies <- (0:(half - 1)) * fs / n
  
  centroid <- sum(frequencies * magnitude) / sum(magnitude)
  
  sqrt(
    sum(magnitude * (frequencies - centroid)^2) /
      sum(magnitude)
  )
}

bandwidth <- apply(
  chunks,
  2,
  spectral_bandwidth,
  fs = fs
)


# spectral entropy

spectral_entropy <- function(z) {
  
  n <- length(z)
  
  fft_z <- fft(z)
  
  half <- floor(n / 2) + 1
  
  magnitude <- Mod(fft_z[1:half])
  
  # Convert to probability distribution
  p <- magnitude / sum(magnitude)
  
  # Avoid log(0)
  p <- p[p > 0]
  
  -sum(p * log(p))
}

entropy <- apply(
  chunks,
  2,
  spectral_entropy
)


# zero crossing rate

zero_crossings <- function(z) {
  
  sum(z[-1] * z[-length(z)] < 0)
}

zcr <- apply(
  chunks,
  2,
  zero_crossings
)

```


## Other time series



```{r}

chunk_time <- ((1:n_chunks) - 1) * chunk_size / fs

plot(
  chunk_time,
  rms,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "RMS amplitude",
  main = "RMS amplitude over time"
)


plot(
  chunk_time,
  max_amp,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "Maximum absolute amplitude",
  main = "Maximum amplitude over time"
)


plot(
  chunk_time,
  sd_amp,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "Standard deviation",
  main = "Amplitude SD over time"
)


plot(
  chunk_time,
  centroid,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "Frequency (Hz)",
  main = "Spectral centroid over time"
)


plot(
  chunk_time,
  bandwidth,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "Bandwidth (Hz)",
  main = "Spectral bandwidth over time"
)

plot(
  chunk_time,
  entropy,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "Spectral entropy",
  main = "Spectral entropy over time"
)

plot(
  chunk_time,
  zcr,
  type = "l",
  xlab = "Time (seconds)",
  ylab = "Zero crossings",
  main = "Zero-crossing rate over time"
)
```


## Other time series



```{r}

features <- data.frame(
  time = chunk_time,
  rms = rms,
  max_amp = max_amp,
  sd = sd_amp,
  centroid = centroid,
  bandwidth = bandwidth,
  entropy = entropy,
  zcr = zcr
)

head(features)


library(forecast)
rms_ts <- ts(rms, frequency = fs / chunk_size)

fit <- auto.arima(rms_ts)

summary(fit)

```


