---
title: "Midterm Project"
author: "Emily Fagerstrom"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
library(tuneR)
library(av)
library(tidyverse)
```


# Cross Correlation


## Stereo Sound: Left and Right

- Two independent time series ($L_t$ and $R_t$) trick your brain 
- Share a lot of information (usually all instruments sent to both)
- Thus, waveforms look close to identical; minor tweaks

## Monocompatibility

- If left and right channels get out of phase, the waves can cancel each other
- If song is played on a smartphone or bluetooth speaker (in mono), sound can vanish
- Correlation analysis to check mixes are "mono-compatible"

## Stereo Tricks

- Don't want the left and right channels to be identical (boring)
- Haas effect: delay a sound in one channel by a few milliseconds
- Idea: look for a lag peak to see the tricks used by the mixing engineer

## Live vs Recorded (vs Remastered)

- Dowload Goo Goo Dolls "Slide" from youtube
- Original
- "Remastered"
- Live in Buffalo July 4, 2004


![](GooGooDollsIris-1){width=50%}

## Fun fact!  44100 Hz

- Limit for healthy hearing: $\approx$ 20 kHz
- Nyquist-Shannon Sampling Them: sample at more than twice the highest frequency you want to record
- Want a little bit of "breathing" room (especially with 1970s analog technology)
- American and European TV were different (30 frames per second vs 25)
- wanted something with a lot of integer factors to cleanly divide audio samples across video frames
- $44,100 = 2^2 \times 3^2 \times 5^2 \times 7^2$
- Sony pitched Compact Disc standard in 1980; locked in now

## CCFs

```{r}
song_og <- readMP3("Slide.mp3")
song_remastered <- readMP3("Slide-Remastered.mp3")
song_live <- readMP3("Slide-Live.mp3")

# Sampling rate (double check it is the usual 44100 Hz)
fs <- song_og@samp.rate
print(paste("Sampling Rate:", fs, "Hz"))

# Tiny smidgin of chorus: "yeaaaah we're gonna let it slide"
start_sec <- 44.2
duration_sec <- 3

# Sample indices
start_sample <- start_sec * fs
end_sample <- (start_sec + duration_sec) * fs

# Left and Right channels
og_left <- song_og@left[start_sample:end_sample]
og_right <- song_og@right[start_sample:end_sample]

#L/R Live version
live_left <- song_live@left[start_sample:end_sample]
live_right <- song_live@right[start_sample:end_sample]

#L/R Remastered version
rem_left <- song_remastered@left[start_sample:end_sample]
rem_right <- song_remastered@right[start_sample:end_sample]




chorus_stereo <- Wave(left = og_left, 
                      right = og_right, 
                      bit = 16, 
                      samp.rate = fs)

writeWave(chorus_stereo, filename = "yeeaaah.wav")


```


<audio controls>
  <source src="yeeaaah.wav" type="audio/wav">
</audio>


## CCFs

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 9
#| out-width: "100%"

par(mfrow = c(3, 1))

ccf_og <- ccf(og_left, og_right, lag.max = 100, 
               main = "CCF: Original (?)",
               ylab = "CCF", col = "blue", lwd = 2)

ccf_rem <- ccf(rem_left, rem_right, lag.max = 100, 
               main = "CCF: Remastered Version",
               ylab = "CCF", col = "red", lwd = 2)

ccf_live <- ccf(live_left, live_right, lag.max = 100, 
                main = "CCF: Live Version",
                ylab = "CCF", col = "green", lwd = 2)

# reset layout
par(mfrow = c(1, 1))

```


## CCFs: Remastered

![](Loudness-wiki){width=75%}

## CCFS: Live Version (Outdoor)

- Smoother: Lots of random continuous background noise
- Negative correlation: peak in the left channel is starting to line up with trough in the right channel
- Lag of 89 relevance:

$\Delta t = \frac{89}{44100} \approx .00202 \,\,\mathrm{seconds}$

- Sound travels through open air at about 1125 feet / second

$1125 \,\, \mathrm{ft/sec} \times .00202 \,\, \mathrm{seconds} \approx 2.27 \,\, \mathrm{feet}$

- The distance between stage mics was about 2.27 feet

- Somebody/something loud was possibly closer to one mic than the other.


##