---
title: "Midterm: Slide Deck"
author: "Emily Fagerstrom"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
library(tuneR)
library(tidyverse)
library(ggplot2)
library(gganimate) #for the animation with the sound
library(forecast)


```


# Cross Correlation

## Stereo Sound: Left and Right

- Two independent time series ($L_t$ and $R_t$) trick your brain 
- Share a lot of information (usually all instruments sent to both)
- Thus, waveforms look close to identical; minor tweaks

## Monocompatibility

- If left and right channels get out of phase, the waves can cancel each other
- If song is played on a smartphone or bluetooth speaker (in mono), sound can vanish
- Correlation analysis to check mixes are "mono-compatible"

## Live vs Recorded (vs Remastered)

- Dowload Goo Goo Dolls "Slide" from youtube
- Original
- "Remastered"
- Live in Buffalo July 4, 2004


![](GooGooDollsIris-1){width=50%}

## Fun fact!  44100 Hz

- Limit for healthy hearing: $\approx$ 20 kHz
- Nyquist-Shannon Sampling Them: sample at more than twice the highest frequency you want to record
- Sound engineers want a little bit of "breathing" room (especially with 1970s analog technology)
- American and European TV were different (30 frames per second vs 25)
- Want a number with a lot of factors to cleanly divide audio samples across video frames
- $44,100 = 2^2 \times 3^2 \times 5^2 \times 7^2$
- Sony pitched Compact Disc standard in 1980; locked in now

## CCFs (left and right channels)

```{r}
song_og <- readMP3("Slide.mp3")
song_remastered <- readMP3("Slide-Remastered.mp3")
song_live <- readMP3("Slide-Live.mp3")

# Sampling rate (double check it is the usual 44100 Hz)
fs <- song_og@samp.rate
print(paste("Sampling Rate:", fs, "Hz"))

# Tiny smidgin of chorus: "yeaaaah we're gonna let it slide"
start_sec <- 44.2
duration_sec <- 3

# Sample indices
start_sample <- start_sec * fs
end_sample <- (start_sec + duration_sec) * fs

# Left and Right channels
og_left <- song_og@left[start_sample:end_sample]
og_right <- song_og@right[start_sample:end_sample]

#L/R Live version
live_left <- song_live@left[start_sample:end_sample]
live_right <- song_live@right[start_sample:end_sample]

#L/R Remastered version
rem_left <- song_remastered@left[start_sample:end_sample]
rem_right <- song_remastered@right[start_sample:end_sample]




chorus_stereo <- Wave(left = og_left, 
                      right = og_right, 
                      bit = 16, 
                      samp.rate = fs)

writeWave(chorus_stereo, filename = "yeeaaah.wav")


```


<audio controls>
  <source src="yeeaaah.wav" type="audio/wav">
</audio>


## CCFs (left and right channels)

```{r}
#| output-location: slide
#| fig-width: 8
#| fig-height: 9
#| out-width: "100%"

par(mfrow = c(3, 1))

ccf_og <- ccf(og_left, og_right, lag.max = 100, 
               main = "CCF: Original (?)",
               ylab = "CCF", col = "blue", lwd = 2)

ccf_rem <- ccf(rem_left, rem_right, lag.max = 100, 
               main = "CCF: Remastered Version",
               ylab = "CCF", col = "red", lwd = 2)

ccf_live <- ccf(live_left, live_right, lag.max = 100, 
                main = "CCF: Live Version",
                ylab = "CCF", col = "green", lwd = 2)

# reset layout
par(mfrow = c(1, 1))

```


## CCFs: Remastered

![](Loudness-wiki){width=75%}


# Autocorrelation

## Note Detection using Rolling Variance

- Intro (first 9 seconds)
- Look at *small* snippets (1024 frames; .23 seconds each)
- Check the variance of each piece
- Places with high variance are likely to be the onset of the next note

## Note Detection using Rolling Variance

```{r}
rm(list = setdiff(ls(), "song_og"))

# audio clip (left channel; intro)
intro_left <- song_og@left[1:(44100 * 9)]
```

```{r}
#| eval: false

# Run this code to create the .mov, but then don't actually run it during the slides, just grab the animation it produced

writeWave(Wave(left = intro_left, samp.rate = 44100, bit = 16), "intro_clip.wav")

# rolling variance dataframe
# (doesn't need to be power of 2 for this purpose;
# but that is usually how things are chunked in audio world to allow for FFTs) 
frame_size <- 1024
n_frames <- floor(length(intro_left) / frame_size)
frames <- matrix(intro_left[1:(n_frames * frame_size)], nrow = frame_size)
frame_vars <- apply(frames, 2, var)
frame_times <- (seq_along(frame_vars) * frame_size) / 44100
df_variance <- data.frame(time = frame_times, variance = frame_vars)

# rolling variance plot and animation
p <- ggplot(df_variance, aes(x = time, y = variance)) +
  geom_line(color = "purple", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Note Onset Detection: Intro Riff",
    subtitle = "Time: {frame_along} seconds",
    x = "Time (Seconds)", 
    y = "Variance (Signal Energy)"
  )

anim <- p + transition_reveal(time)

# combine animation and audio
animate(anim, 
        nframes = 180, 
        fps = 20, 
        renderer = av_renderer(file = "variance_video.mp4", audio = "intro_clip.wav"))

```


## Note Detection using Rolling Variance

{{< video variance_video.mp4 width="700" >}}

## Pitch Detection using ACF

First note of the song (.7 to .9 sec)

```{r}
# Starting note
note_start_sec <- .7
note_end_sec <- .9

note_segment <- intro_left[(note_start_sec * 44100):(note_end_sec * 44100)]
writeWave(Wave(left = note_segment, samp.rate = 44100, bit = 16), "first_note.wav")

# ACF
acf(note_segment, lag.max = 300, 
    main = "ACF (a single note)",
    col = "blue", lwd = 2)

# PACF
pacf(note_segment, lag.max = 300, 
    main = "PACF (a single note)",
    col = "blue", lwd = 2)


```

## Pitch Detection using ACF
- At lag 251, autocorrelation is almost 1
- $44100/250 \approx 175.697 \approx F$


```{r}
#| eval: false


freq <- 175.7
duration <- 1
sample_rate <- 44100

t <- seq(0, duration, 1 / sample_rate)
wave <- sin(2 * pi * freq * t)

wav <- Wave(
  left = as.integer(wave * 32767),
  samp.rate = sample_rate,
  bit = 16
)

writeWave(wav, "176Hz.wav")

```

The intro (left channel):

<audio controls>
  <source src="intro_clip.wav" type="audio/wav">
</audio>

The first note (.7 to .9 sec):

<audio controls>
  <source src="first_note.wav" type="audio/wav">
</audio>

A 175.7 Hz sound wave:

<audio controls>
  <source src="176Hz.wav" type="audio/wav">
</audio>



```{r}
plot(note_segment, type="l")

spec.pgram(note_segment,
           Fs = 44100,
           log = "no")


fit <- Arima(note_segment,
             order = c(20, 0, 0),
             include.mean = TRUE)

summary(fit)


ar_coefs <- coef(fit)[grep("^ar", names(coef(fit)))]

```




