---
title: "HW 1 - 9/10"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
library(forecast)
library(fpp3)
library(tidyverse)
library(tsbox)
library(zoo)
library(seasonal)
library(astsa)
```

## Gas and Unemployment

Looking at gas prices with unemployment claims.  Here's the last 5 years


```{r}
library(fredr)
fredr_set_key("052142bc981666b4ebcb1c8df98d006b")

icnsa = fredr(series_id = "ICNSA")
gasregw = fredr(series_id = "GASREGW")

icnsa <- icnsa |>
  mutate(date = as.Date(date)) |>
  arrange(date)

gasregw <- gasregw |>
  mutate(date = as.Date(date)) |>
  arrange(date)

lastfive <- tail(icnsa,260)

lastfive |> 
  select(time = date, value) |>
  ts_dygraphs()


last5gas <- tail(gasregw,260)


last5gas |> 
  select(time = date, value) |>
  ts_dygraphs()
```


## Adjustments?

Unemployment claims doesn't seem to be adjusted for population (e.g. "per x thousand" or whatever), but also doesn't show upward trend? I'm surprised.

Gas should probably be adjusted to some year's dollars?

but... too bad.  not going to

```{r}
gasregw |> 
  select(time = date, value) |>
  ts_dygraphs()


```


## Lags

Huge lag for weeks 1-5.  (That is, data from 1-5 weeks ago seems to tell you a lot about current claims.)  Not a huge spike at 52, which is surprising?

```{r}
icnsa_ts <- icnsa |>
  as_tsibble(index = date)

icnsa_ts |>
  ACF(value, lag_max = 60) |>
  autoplot()

```

## Difference

ok not actually sure what to make of this one.

```{r}
icnsa_ts |>
  autoplot(difference(value)) +
  labs(
    title = "Weekly Change in ICNSA",
    y = "Change in claims",
    x = "Week"
  )

icnsa_ts |>
  ACF(difference(value), lag_max = 20) |>
  autoplot()


```


## Looking at lag 1 more


```{r}
icnsa_ts |>
  ggplot(aes(x = lag(value, 1), y = value)) +
  geom_point() +
  labs(
    title = "ICNSA: Current Week vs Previous Week",
    x = "ICNSA at lag 1",
    y = "ICNSA this week"
  )


fit_ar1 <- icnsa_ts |>
  model(ARIMA(value ~ pdq(1,0,0)))

report(fit_ar1)

fit_ar1 |>
  forecast(h = 1)
```

So if we say

$$X_t = c + \phi_1 X_{t-1} + Z_t$$

then estimate is 184236


## Seasonal, so ... force a lag 52?

Expect to see spikes at multiples of 52.
But... not really within the blue bars.

Model would be

$$
X_t = \beta_0 + \beta_1 X_{t-1} + \beta_2 X_{t-52} + \epsilon_t
$$
Next prediction would be:
181484.8


```{r}
icnsa_ts |>
  ACF(value, lag_max = 52) |>
  autoplot()

icnsa_ts |>
  ACF(value, lag_max = 260) |>
  autoplot()



icnsa_lag <- icnsa |>
  arrange(date) |>
  mutate(
    lag1 = lag(value, 1),
    lag52 = lag(value, 52)
  )


model <- lm(value ~ lag1 + lag52, data = icnsa_lag)

summary(model)

new_data <- data.frame(
  lag1 = 170626,
  lag52 = 204862
)

predict(model, newdata = new_data)
```


## Bring in gas prices

not going to worry about real vs nominal dollars, even though I think that would be the right move

Data for gas starts in 1990; claims data starts in 1967
Also, the data comes out on a different day of the week

```{r}

gasregw_ts <- gasregw |>
  as_tsibble(index = date)

gasregw_ts |>
  autoplot(value) +
  labs(
    title = "Weekly Regular Gasoline Prices",
    y = "Dollars per gallon",
    x = "Week"
  )

range(icnsa$date)
range(gasregw$date)

```


## Alignment. Bad

This was not the best way to do this. In particular, one of these datasets is missing a data point, because the weekly alignment gets off/mismatched.

Too bad.  Close enough.

Looks like, when gas prices are low, unemployment claims are high, and vice versa.  I'm going to force gas into the model for homework reasons.  And at lag 0.

This isn't quite right... we wouldn't "know ahead of time" the gas price any more than we would know the claims.  So I guess this is a lag 1.  (But also, the day of the week that these come out is offset, so...)


Also, now that gas is in the model, lag52 is no longer statistically significant.  I think the model is better without gas.  BUT!  prediction is 166248.

```{r}

icnsa_gas <- data.frame(
  icnsa = tail(icnsa$value, 1000),
  gas   = tail(gasregw$value, 1000)
)


ccf(icnsa_gas$gas, icnsa_gas$icnsa,
    lag.max = 60,
    main = "CCF: Gas Prices and Initial Claims")

icnsa_gas_lag <- icnsa_gas |>
  mutate(
    lag1 = lag(icnsa, 1),
    lag52 = lag(icnsa, 52)
  )


model_gas <- lm(
  icnsa ~ lag1 + lag52 + gas,
  data = icnsa_gas_lag
)

summary(model_gas)


new_data <- data.frame(
  lag1 = icnsa_gas_lag$icnsa[nrow(icnsa_gas_lag)],
  lag52 = icnsa_gas_lag$icnsa[nrow(icnsa_gas_lag) - 51],
  gas = icnsa_gas_lag$gas[nrow(icnsa_gas_lag)]
)

predict(model_gas, newdata = new_data)
```
