---
title: "Homework 1 Solution: Forecasting Initial Claims (ICNSA)"
subtitle: "UB Time Series -- Fall 2026"
author: "Jim Livsey"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 8
    fig-height: 4.5
    theme: cosmo
  gfm:
    toc: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Overview

This document works through Homework 1 (see `hw/hw1/hw1.md`):

1. Load the most current **Initial Claims (ICNSA)** series (not seasonally adjusted).
2. Choose and load a related covariate series.
3. Build a **regression-based** forecast of ICNSA for the week ending **September 5, 2026**.
4. Build an **ARIMA** forecast for the same week, using a seasonal period of **52** weeks.
5. Use empirical analysis (ACF/PACF, unit-root tests, information criteria, residual
   diagnostics) to choose the ARIMA model's stochastic structure, following the classic
   **Box--Jenkins** identify -> estimate -> diagnose cycle.
6. Report both forecasts as the final two outputs.

At every step the code is accompanied by a discussion of *why* a particular choice was
made, not just *what* was done, that reasoning is the actual point of the exercise.

```{r}
#| label: setup
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forecast)
library(tseries)
library(scales)

theme_set(theme_minimal(base_size = 11))
```

## Step 1: Load the most current ICNSA data

FRED's `ICNSA` series ("Initial Claims", not seasonally adjusted, weekly, ending
Saturday) 

```r
fred_url <- "https://fred.stlouisfed.org/graph/fredgraph.csv?id=ICNSA"
icnsa_raw <- read.csv(fred_url)
```

For reproducability the data is stored in `data/ICNSA.csv`.
FRED revises NSA claims data
  occasionally and the series grows every week, so re-downloading later would silently
  change the training data and the reported forecasts would no longer match this
  write-up. 

```{r}
#| label: load-icnsa
icnsa <- read_csv("data/ICNSA.csv", show_col_types = FALSE) |>
  rename(date = DATE, icnsa = VALUE) |>
  arrange(date)

cat("ICNSA range:", format(min(icnsa$date)), "to", format(max(icnsa$date)),
    "(", nrow(icnsa), "weekly observations )\n")
tail(icnsa, 5)
```

```{r}
#| label: fig-icnsa-full
#| fig-cap: "Initial Claims (ICNSA), full history"
ggplot(icnsa, aes(date, icnsa)) +
  geom_line(linewidth = 0.3) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = NULL, y = "Initial claims (NSA)",
       title = "ICNSA, 1967-present")
```

Two features of the full history matter for every modeling choice that follows:

* **Strong, stable annual seasonality** superimposed on a much slower-moving level --
  claims spike every December/January (holiday layoffs, seasonal hiring cycles) and
  trough in the spring/summer.
* **One enormous, short-lived anomaly**: the March-June 2020 COVID-19 shock, where
  weekly claims jumped from a normal range of roughly 200,000-400,000 to over
  **6.16 million** in a single week (2020-04-04) before decaying back down over about a
  year. This single event is orders of magnitude larger than any ordinary business-cycle
  fluctuation in the other 59 years of data, and if left untreated it will dominate
  variance estimates, unit-root tests, and autocorrelation estimates for *any* model fit
  to the raw series. It has to be dealt with explicitly rather than ignored -- see below.

### Choosing a modeling window

Rather than using the full 1967-2026 history, the regression and ARIMA models below are
both fit on data from **2010-01-02 onward**. The reasoning:

* The UI claims system, benefit rules, and the structure of the labor market itself have
  changed enough since the 1970s that pooling 59 years of data would mix several
  different data-generating regimes -- not a good idea for a single seasonal ARIMA
  model.
* Sixteen-plus years (869 weekly observations) is still far more than needed to
  estimate a seasonal model with period 52 reliably (a common rule of thumb is at least
  4-5 full seasonal cycles; this window has more than 16).
* It comfortably includes a full business cycle (the 2020 COVID recession and
  recovery), which is useful for the model to have seen, *provided* that episode is
  explicitly flagged rather than treated as ordinary noise.

```{r}
#| label: window
win_start   <- as.Date("2010-01-02")
target_date <- as.Date("2026-09-05")   # week we are forecasting

icnsa_win <- icnsa |> filter(date >= win_start, date < target_date)
cat("Modeling window:", format(min(icnsa_win$date)), "to", format(max(icnsa_win$date)),
    "(", nrow(icnsa_win), "observations )\n")
```

### Handling the COVID-19 anomaly

Because the 2020 shock is so extreme (and known, dated, and one-time), the cleanest fix
is an explicit **intervention dummy** rather than trying to make a purely stochastic
model absorb it. The window below was chosen empirically by looking at which weeks
exceed the highest *ordinary* seasonal peak seen anywhere else in the post-2010 window
(about 900,000, the typical January maximum):

```{r}
#| label: covid-window
icnsa_win |> filter(icnsa > 900000) |> select(date, icnsa) |> as.data.frame()
```

Claims first jump above that threshold on 2020-03-21 and the last observation still
visibly elevated above the normal seasonal envelope is in mid-February 2021. To give a
little buffer on both ends, the intervention window is set to **2020-03-14 through
2021-03-31**, encoded as a single 0/1 dummy variable used as a regressor in both the
regression model and (as `xreg`) in the ARIMA model below. A single step dummy is a
simplification -- the true shape is a sharp spike followed by a decay, not a clean
rectangle -- but it removes the overwhelming majority of the distortion with one
parameter, which is the right amount of complexity for this assignment; the residual
diagnostics later on show exactly where this simplification leaves its mark.

```{r}
#| label: covid-dummy
icnsa_win <- icnsa_win |>
  mutate(covid = as.integer(date >= as.Date("2020-03-14") & date <= as.Date("2021-03-31")))
```

## Step 2: A covariate for ICNSA

**Continued Claims (Insured Unemployment)**, FRED series `CCNSA`, is the natural choice
of covariate:

* It is published in the *same* weekly Unemployment Insurance Claims release as ICNSA,
  by the same source, on the same NSA basis -- so there is no unit- or
  methodology-mismatch to worry about.
* Economically, continued claims are mechanically related to initial claims: people who
  file new (initial) claims and remain unemployed show up as continued claims in
  subsequent weeks, so the two series should be strongly cross-correlated.
* Practically, it is genuinely useful as a *predictor* rather than just a correlated
  series, because of a real-time data-availability quirk described next.

```{r}
#| label: load-ccnsa
ccnsa <- read_csv("data/CCNSA.csv", show_col_types = FALSE) |>
  rename(date = DATE, ccnsa = VALUE) |>
  arrange(date)

cat("CCNSA range:", format(min(ccnsa$date)), "to", format(max(ccnsa$date)),
    "(", nrow(ccnsa), "weekly observations )\n")
tail(ccnsa, 3)
```

Notice that the most recent CCNSA observation is for the week ending **2026-08-22**,
one week *behind* the most recent ICNSA observation (2026-08-29). This is not a data
error -- continued claims for a given week cannot be finalized until people have had a
chance to *continue* claiming into the following week, so CCNSA is always released one
week later in "event time" than ICNSA. That has a direct consequence for forecasting:
by the time we stand at 2026-09-05 and want to forecast, the newest CCNSA value we could
actually have in hand corresponds to the week ending 2026-08-22 -- two weeks before our
target date, not one.

### Choosing the lag empirically

To decide *how* to use CCNSA rather than just *whether* to, look at the
cross-correlation function between weekly changes (first differences of the logs, to
remove the shared trend/level and put both series on a comparable multiplicative
scale) in the two series:

```{r}
#| label: fig-ccf
#| fig-cap: "CCF between weekly log-changes in CCNSA and ICNSA"
cc_df <- icnsa_win |>
  left_join(ccnsa, by = "date") |>
  arrange(date) |>
  mutate(dlogI = c(NA, diff(log(icnsa))),
         dlogC = c(NA, diff(log(ccnsa)))) |>
  filter(!is.na(dlogI), !is.na(dlogC))

ccf_obj <- ccf(cc_df$dlogC, cc_df$dlogI, lag.max = 6, plot = TRUE,
               main = "d(log CCNSA) vs d(log ICNSA)")
data.frame(lag = ccf_obj$lag, correlation = round(ccf_obj$acf, 3))
```

Reading this CCF (R's convention: the value at `lag = k` is the correlation between
`CCNSA` shifted forward by `k` weeks and contemporaneous `ICNSA`):

* The strongest correlation is at **lag +1** (about 0.52): *this* week's initial claims
  changes predict *next* week's continued-claims changes -- which makes complete sense
  causally (today's new filers become next week's continuers) but is useless for
  forecasting ICNSA, since it runs in the wrong direction.
* Among the lags that actually run the *useful* direction (CCNSA change leading ICNSA
  change), the strongest is **lag -1** (about 0.37): last week's CCNSA change helps
  predict this week's ICNSA change. Unfortunately, as established above, CCNSA at
  `lag -1` relative to our target date (2026-09-05) would be the week ending 2026-08-29
  -- which is **not yet published**.
* The strongest correlation at a lag that is *actually available in real time* given
  the one-week publication delay is **lag -2** (about -0.22): a smaller, negative, but
  still non-trivial relationship.

So there is a real tension here between the statistically strongest lag relationship
and the one that is operationally usable -- a distinction that matters a lot in
real-time forecasting and is worth being explicit about rather than silently picking
whichever lag happens to have the best in-sample fit. The regression below uses
**CCNSA lagged 2 weeks** because that is the information actually in hand on the
forecast date; the negative sign that shows up in the fitted model is a genuine, if
modest, empirical relationship in this window, not a sign error.

```{r}
#| label: lag-covariate
icnsa_win <- icnsa_win |>
  left_join(ccnsa |> rename(ccnsa_val = ccnsa), by = "date") |>
  arrange(date) |>
  mutate(logICNSA = log(icnsa),
         logCCNSA_lag2 = lag(log(ccnsa_val), 2))
```

## Step 3: Regression-based forecast

### Model specification

Three features of ICNSA drive the choice of regression specification:

* **Multiplicative variation.** The size of both the trend and the seasonal swings
  scales with the level of claims (seasonal troughs and peaks are much bigger in
  absolute terms when the base level is high). Modeling `log(ICNSA)` rather than
  `ICNSA` turns this multiplicative structure into an additive one, which linear
  regression handles far better and also guarantees back-transformed forecasts are
  positive.
* **A slow, roughly linear drift** in the log level over the window, captured with a
  simple linear time trend `t` (in weeks since the start of the window).
* **Annual seasonality whose period does not divide evenly into the data.** A calendar
  year is about 365.25/7 = 52.18 weeks, not an integer, so `factor(week-of-year)`
  dummies do not stay aligned with the calendar over 16+ years (the misalignment
  compounds by roughly a week every six years). The standard regression-based fix is a
  small number of Fourier (harmonic) pairs at the true period `m = 365.25/7`, which
  handles the non-integer period exactly and cleanly. Three harmonic pairs (`K = 3`)
  captures the main annual shape (a strong first harmonic plus corrections) without
  overfitting high-frequency noise.

Putting these together, plus the COVID dummy and the lag-2 CCNSA covariate from Step 2:

$$
\log(\text{ICNSA}_t) = \beta_0 + \beta_1 t + \beta_2\,\text{covid}_t +
\beta_3 \log(\text{CCNSA}_{t-2}) +
\sum_{k=1}^{3}\Big[a_k \sin\Big(\tfrac{2\pi k t}{m}\Big) + b_k \cos\Big(\tfrac{2\pi k t}{m}\Big)\Big] + \varepsilon_t,
\qquad m = \tfrac{365.25}{7}
$$

```{r}
#| label: fit-regression
reg_data <- icnsa_win |> filter(!is.na(logCCNSA_lag2))
t0 <- min(reg_data$date)
reg_data <- reg_data |> mutate(t = as.numeric(date - t0) / 7)

m <- 365.25 / 7
K <- 3
make_fourier <- function(t, K, m) {
  X <- matrix(NA_real_, nrow = length(t), ncol = 2 * K)
  for (k in seq_len(K)) {
    X[, 2 * k - 1] <- sin(2 * pi * k * t / m)
    X[, 2 * k]     <- cos(2 * pi * k * t / m)
  }
  colnames(X) <- paste0(rep(c("sin", "cos"), K), rep(seq_len(K), each = 2))
  as.data.frame(X)
}

fourier_train <- make_fourier(reg_data$t, K, m)
model_df <- cbind(logICNSA = reg_data$logICNSA, t = reg_data$t,
                   covid = reg_data$covid, logCCNSA_lag2 = reg_data$logCCNSA_lag2,
                   fourier_train)

reg_fit <- lm(logICNSA ~ ., data = model_df)
summary(reg_fit)
```

Every term is highly significant, the COVID dummy and the CCNSA lag both have the
expected sign, and the fit ($R^2 \approx 0.82$) is a large improvement over a
trend + seasonality model with no covariate (adjusted $R^2 \approx 0.16$ for a
harmonic-only model on this series, using the same technique as the Homework 0
solution) -- direct evidence that continued claims carry real, useful information
about initial claims beyond just the shared trend and seasonal cycle.

### Regression residual diagnostics

```{r}
#| label: fig-reg-resid
#| fig-cap: "Regression residuals over time"
plot(reg_data$date, residuals(reg_fit), type = "l",
     xlab = NULL, ylab = "residual (log scale)",
     main = "OLS regression residuals")
abline(h = 0, col = "grey50", lty = 2)
```

```{r}
#| label: reg-ljungbox
Box.test(residuals(reg_fit), lag = 52, type = "Ljung-Box", fitdf = length(coef(reg_fit)) - 1)
```

The Ljung-Box test overwhelmingly rejects the null of no residual autocorrelation. This
is expected and important: ordinary least squares only models the *mean* structure
(trend, seasonality, covariate effect); it says nothing about the fact that a week with
unusually high claims tends to be followed by other weeks with unusually high claims.
That leftover serial dependence is exactly what an ARIMA error structure is designed to
capture -- which motivates Step 4.

### Regression forecast

```{r}
#| label: regression-forecast
t_target <- as.numeric(target_date - t0) / 7
last_logCCNSA <- log(ccnsa$ccnsa[ccnsa$date == target_date - 14])
fourier_target <- make_fourier(t_target, K, m)
newdata <- cbind(t = t_target, covid = 0, logCCNSA_lag2 = last_logCCNSA, fourier_target)

reg_pred_log <- predict(reg_fit, newdata = newdata, interval = "prediction", level = 0.95)
reg_pred <- exp(reg_pred_log)
reg_forecast <- unname(reg_pred[1, "fit"])

cat(sprintf("Regression forecast for the week ending %s: %s  (95%% PI: %s - %s)\n",
            format(target_date, "%B %d, %Y"),
            comma(round(reg_forecast)),
            comma(round(reg_pred[1, "lwr"])),
            comma(round(reg_pred[1, "upr"]))))
```

## Step 4-5: ARIMA forecast via the Box-Jenkins methodology

The homework asks for a seasonal ARIMA fit using **frequency 52** for the seasonal
component, and for the stochastic structure to be chosen empirically. The classic
Box-Jenkins recipe -- **identify, estimate, diagnose (repeat as needed)** -- is followed
explicitly below, working with the same log-transformed, COVID-adjusted series used
above (the log transform is retained for the same variance-stabilization reason as in
Step 3; the COVID period is handled the same way, as a regressor rather than by
deleting data, so that the seasonal ARIMA structure is estimated from the whole
16-year window).

```{r}
#| label: arima-setup
# The ARIMA step is univariate (no covariate needed), so it uses the full
# 2010-2026 ICNSA window directly rather than the 2-observations-shorter
# data frame used for the regression (which loses its first two rows to
# the lag-2 CCNSA covariate).
y <- ts(icnsa_win$logICNSA, frequency = 52, start = c(2010, 1))
xreg_covid <- matrix(icnsa_win$covid, ncol = 1, dimnames = list(NULL, "covid"))
```

A note on `frequency = 52`: a calendar year is actually about 52.18 weeks, so a fixed
52-week seasonal period is a (standard, necessary) approximation -- `ts()` requires an
integer frequency, and 52 is the conventional choice for weekly economic data. The
practical consequence is that the model's notion of "week 1 of the seasonal cycle"
slowly drifts relative to the actual calendar (and relative to moving holidays like
Thanksgiving) by about a week every six years. That drift shows up later as a small,
structured pattern in the residuals -- worth flagging now, revisited in the diagnostics.

### Identification

**(a) Is the series stationary in levels?** Two classroom-standard tests disagree,
which is itself informative:

```{r}
#| label: unit-root-level
adf.test(y)
kpss.test(y, null = "Level")
```

The ADF test (null = unit root) rejects at the 5% level, suggesting stationarity, while
the KPSS test (null = stationarity) also rejects, suggesting *non*-stationarity. This
contradiction is a well-known symptom of a series with strong seasonal structure: both
tests are built around a non-seasonal alternative and can behave erratically when a
large seasonal component is present. Formal automated seasonal-differencing tests give
a similarly unhelpful signal:

```{r}
#| label: nsdiffs
nsdiffs(y)
ndiffs(y)
```

`nsdiffs()` reports that *no* seasonal differencing is needed, which conflicts with
the very obvious seasonal cycle in @fig-icnsa-full. Rather than trust an automated test
in a situation where it is known to be unreliable, the classic Box-Jenkins approach --
look directly at the ACF -- is more trustworthy here.

**(b) Visual identification via the ACF at seasonal lags.** First, the COVID dummy is
regressed out (on its own) purely to get a cleaner look at the underlying seasonal
autocorrelation without the pandemic spike contaminating it:

```{r}
#| label: acf-seasonal-lags
r <- residuals(lm(logICNSA ~ covid, data = icnsa_win))
r_ts <- ts(r, frequency = 52)

acf_r <- Acf(r_ts, lag.max = 4 * 52, plot = FALSE)
seas_lags <- seq(52, 4 * 52, by = 52)
data.frame(lag = seas_lags, acf = round(acf_r$acf[seas_lags + 1], 3))
```

The autocorrelation at exact multiples of 52 weeks decays slowly (0.58, 0.34, 0.25,
0.16 at 1, 2, 3, 4 years) rather than cutting off -- the textbook signature of a
**seasonal unit root**, i.e. evidence *for* seasonal differencing (`D = 1`) even though
the automated `nsdiffs()` test above said otherwise.

```{r}
#| label: seasonal-diff-check
r_sd <- diff(r_ts, lag = 52)
adf.test(r_sd)
Acf(r_sd, lag.max = 10, plot = FALSE)$acf[1:6]
```

After one seasonal difference, the ADF test rejects a unit root -- but the ACF at
short lags (0.78, 0.65, 0.60, 0.54, ...) still decays very slowly, which is the
signature of a remaining **non-seasonal** unit root, so a regular first difference is
also warranted:

```{r}
#| label: regular-diff-check
r_sd_d <- diff(r_sd, lag = 1)
adf.test(r_sd_d)
cat("Variance before regular differencing:", round(var(r_sd), 5),
    " after:", round(var(r_sd_d), 5), "\n")
```

The ADF test now rejects strongly, and critically the variance *drops* (0.075 to
0.034) rather than increasing after differencing -- if the series had already been
stationary, differencing again would have inflated the variance (over-differencing),
so this drop confirms the extra difference is warranted rather than harmful. This
gives `d = 1`, `D = 1` at `s = 52` -- so the working series for identifying the ARMA
orders is $w_t = (1-B)(1-B^{52})r_t$.

**(c) Identifying `(p,q)` and `(P,Q)` from the ACF/PACF of `w_t`.**

```{r}
#| label: fig-acf-pacf-final
#| fig-cap: "ACF and PACF of the fully differenced, COVID-adjusted series"
par(mfrow = c(2, 1))
Acf(r_sd_d, lag.max = 60, main = "ACF")
Pacf(r_sd_d, lag.max = 60, main = "PACF")
```

Two patterns stand out, each read the classic Box-Jenkins way:

* **Non-seasonal lags (1-2):** both the ACF and PACF have significant spikes at lags 1
  and 2 and are essentially inside the confidence band from lag 3 onward. A pattern
  that "cuts off" in *both* ACF and PACF after a couple of lags, rather than one of
  them decaying smoothly, points toward a low-order mixed `ARMA(p,q)` for the
  non-seasonal part rather than a pure AR or pure MA -- so the search below considers
  small combinations of `p` and `q` up to 2 instead of committing to one shape a priori.
* **Seasonal lag (52):** there is one large, isolated spike at exactly lag 52 in both
  the ACF (about -0.43) and PACF (about -0.45), with essentially nothing at the
  neighboring lags (48-51, 53-56). A single seasonal spike that then vanishes is the
  textbook signature of a seasonal `MA(1)` (or, less commonly, `AR(1)`) at `s = 52` --
  so the seasonal order search below focuses on `P, Q \in \{0, 1\}`.

### Estimation: comparing candidate models

Rather than an exhaustive grid search over every combination (computationally wasteful
and not how Box-Jenkins is meant to be practiced), the identification step above
narrows the field to a small number of substantively motivated candidates, each fit
with the COVID dummy as an exogenous regressor (`xreg`) so the intervention does not
contaminate the ARMA parameter estimates:

```{r}
#| label: candidate-models
#| cache: true
candidates <- list(
  c(1, 1, 1,  0, 1, 1),
  c(0, 1, 2,  0, 1, 1),
  c(2, 1, 0,  0, 1, 1),
  c(0, 1, 1,  1, 1, 0),
  c(2, 1, 1,  0, 1, 1),
  c(1, 1, 2,  0, 1, 1)
)

fit_one <- function(cc) {
  tryCatch(
    Arima(y, order = c(cc[1], cc[2], cc[3]),
          seasonal = list(order = c(cc[4], cc[5], cc[6]), period = 52),
          xreg = xreg_covid, method = "CSS-ML"),
    error = function(e) NULL)
}

fits <- lapply(candidates, fit_one)
keep <- !sapply(fits, is.null)
candidates <- candidates[keep]
fits <- fits[keep]

candidate_table <- do.call(rbind, Map(function(cc, fit) {
  lb <- Box.test(residuals(fit), lag = 104,
                  fitdf = sum(cc[c(1, 3, 4, 6)]), type = "Ljung-Box")
  data.frame(
    model = sprintf("(%d,%d,%d)(%d,%d,%d)[52]", cc[1], cc[2], cc[3], cc[4], cc[5], cc[6]),
    AICc = round(fit$aicc, 2),
    BIC  = round(fit$bic, 2),
    ljung_box_p = round(lb$p.value, 3)
  )
}, candidates, fits))

candidate_table
```

`(1,1,2)(0,1,1)[52]` has the lowest AICc *and* the largest Ljung-Box p-value (i.e. the
residuals look most like white noise) among the candidates considered, and its
coefficients are all comfortably significant:

```{r}
#| label: final-model
best_fit <- fits[[which.min(sapply(fits, function(f) f$aicc))]]
best_fit
```

### Diagnostic checking

```{r}
#| label: fig-residual-diagnostics
#| fig-cap: "Residual diagnostics for the selected SARIMA model"
checkresiduals(best_fit)
```

```{r}
#| label: ljung-box-final
Box.test(residuals(best_fit), lag = 104, fitdf = 4, type = "Ljung-Box")
shapiro.test(as.numeric(residuals(best_fit)))
```

Two things to take away:

* **Autocorrelation: clean.** The Ljung-Box test does not reject at conventional
  levels, and the residual ACF in the plot above has no meaningful spikes -- the
  `(1,1,2)(0,1,1)[52]` structure has captured essentially all of the linear serial
  dependence in the series, which is the primary thing a Box-Jenkins diagnostic check
  is trying to establish.
* **Normality: fails, for an explainable reason.** The Shapiro-Wilk test strongly
  rejects normality, and the residual histogram has heavy tails. Looking at the largest
  residuals by date:

```{r}
#| label: residual-outliers
icnsa_win |>
  mutate(resid = as.numeric(residuals(best_fit))) |>
  arrange(desc(abs(resid))) |>
  select(date, icnsa, resid) |>
  head(8) |>
  as.data.frame()
```

  the single largest residual is the very first week of the COVID spike
  (2020-03-21) -- a single step dummy cannot fully capture a shock that actually
  arrives as a sharp spike-and-decay rather than a clean rectangle, so a little of that
  shape leaks into the residuals right at the onset and offset of the intervention
  window. Beyond that, several of the next-largest residuals recur every
  **late November / early December** (2013, 2019, 2024) -- almost certainly a
  **Thanksgiving effect**: the holiday moves around within the calendar month from year
  to year, so its impact does not land on the same "week of the 52-week seasonal cycle"
  every year, which is exactly the `frequency = 52` approximation flagged earlier. Both
  effects are known, dated, and small relative to the overall signal; they do not
  indicate that the model has missed real linear dependence (Ljung-Box is clean), only
  that its errors are not perfectly Gaussian -- so prediction intervals below should be
  read as reasonable but somewhat optimistic in the tails.

### ARIMA forecast

```{r}
#| label: arima-forecast
arima_fc <- forecast(best_fit, h = 1,
                      xreg = matrix(0, nrow = 1, ncol = 1, dimnames = list(NULL, "covid")),
                      level = 95)
arima_forecast <- exp(as.numeric(arima_fc$mean))
arima_lo <- exp(arima_fc$lower[1, 1])
arima_hi <- exp(arima_fc$upper[1, 1])

cat(sprintf("SARIMA(1,1,2)(0,1,1)[52] forecast for the week ending %s: %s  (95%% PI: %s - %s)\n",
            format(target_date, "%B %d, %Y"),
            comma(round(arima_forecast)), comma(round(arima_lo)), comma(round(arima_hi))))
```

## Step 6: Final forecasts

Both approaches -- a harmonic regression with a lagged continued-claims covariate, and
a seasonal ARIMA model identified through the classic Box-Jenkins cycle -- are fit
independently and land within about 4% of each other, which is a reassuring
cross-check given how differently the two methods handle serial dependence: the
regression relies on an external covariate and a smooth deterministic seasonal shape,
while the ARIMA model relies entirely on the series' own past values and errors, with
seasonality captured stochastically.

```{r}
#| label: final-summary
final_table <- data.frame(
  Method = c("Regression (trend + Fourier + COVID dummy + CCNSA lag-2)",
             "SARIMA(1,1,2)(0,1,1)[52] with COVID xreg"),
  Forecast = comma(round(c(reg_forecast, arima_forecast))),
  `95% PI` = c(paste0(comma(round(reg_pred[1, "lwr"])), " - ", comma(round(reg_pred[1, "upr"]))),
               paste0(comma(round(arima_lo)), " - ", comma(round(arima_hi)))),
  check.names = FALSE
)
knitr::kable(final_table,
             caption = paste("ICNSA forecasts for the week ending", format(target_date, "%B %d, %Y")))
```

```{r}
#| label: forecast-1-regression
reg_forecast
```

```{r}
#| label: forecast-2-arima
arima_forecast
```
