---
title: "hw1_dimmig"
format: html
editor: visual
---

```{r}
library(fredr)
library(tidyverse)
library(tsibble)
library(fable)
library(feasts)
library(fabletools)
library(forecast)     
library(tsbox)        
library(zoo)
library(tseries)     
library(lubridate)

fred_url <- "https://fred.stlouisfed.org/graph/fredgraph.csv?id=ICNSA"

icnsa_raw <- read.csv(fred_url, stringsAsFactors = FALSE)
names(icnsa_raw) <- c("date", "value")

icnsa <- data.frame(
  date  = as.Date(icnsa_raw$date),
  ICNSA = suppressWarnings(as.numeric(icnsa_raw$value))
)

print(tail(icnsa, 5))

win_start   <- as.Date("2010-01-02")
target_date <- as.Date("2026-09-05")   # week we are forecasting

icnsa_win <- icnsa |> filter(date >= win_start, date < target_date)
cat("Modeling window:", format(min(icnsa_win$date)), "to", format(max(icnsa_win$date)),
    "(", nrow(icnsa_win), "observations )\n")

icnsa_win <- icnsa_win |>
  mutate(covid = as.integer(date >= as.Date("2020-03-14") & date <= as.Date("2021-03-31")))

# adding in the covariate of CCNSA
ccnsa_raw <- read_csv(
  "https://fred.stlouisfed.org/graph/fredgraph.csv?id=CCNSA",
  show_col_types = FALSE
) |>
  rename(date = observation_date, CCNSA = CCNSA) |>
  mutate(date = as.Date(date)) |>
  filter(!is.na(CCNSA))

dat <- icnsa_win |>
  inner_join(ccnsa_raw, by = "date") |>
  mutate(week = yearweek(date)) |>
  as_tsibble(index = week) |>
  mutate(log_ICNSA = log(ICNSA),
         log_CCNSA = log(CCNSA))

autoplot(dat, ICNSA) +
  labs(title = "Initial Claims (ICNSA), weekly, NSA", y = "Claims")

autoplot(dat, log_ICNSA) +
  labs(title = "log(ICNSA)", y = "log(Claims)")

dlog_ICNSA <- diff(dat$log_ICNSA)
dlog_CCNSA <- diff(dat$log_CCNSA)

ccf(dlog_ICNSA,dlog_CCNSA,10)

icnsa_win <- icnsa_win |> left_join(ccnsa_raw) |> arrange(date) |> mutate(logICNSA = log(ICNSA), logCCNSA_lag2 = lag(log(CCNSA),2))
reg_data <- icnsa_win |> filter(!is.na(logCCNSA_lag2))
model_df2 <- as.data.frame(cbind(logICNSA = reg_data$logICNSA, t = reg_data$t,
                                 covid = reg_data$covid, logCCNSA_lag2 = reg_data$logCCNSA_lag2))
reg_fit2 <- lm(logICNSA ~ ., data = model_df2)
summary(reg_fit2)

t0 <- min(reg_data$date)
t_target <- as.numeric(target_date - t0)/7
last_logCCNSA <- log(ccnsa_raw$CCNSA[ccnsa_raw$date == target_date - 14])
sep_5 <- as.data.frame(cbind(t = t_target, covid = 0, logCCNSA_lag2 = last_logCCNSA))



```

Predicting using an ARIMA model:

```{r}
y <- ts(icnsa_win$logICNSA, frequency = 52, start = c(2010, 1))
xreg_covid <- matrix(icnsa_win$covid, ncol = 1, dimnames = list(NULL, "covid"))

ts_plot(y)
ts_plot(diff(y))


```

Based on the original plot of the log(ICNSA) values, we can see a trend and non seasonality. When taking the differential, the data becomes stationary. Therefore, d = 1.

```{r}

d_order <- ndiffs(y)
D_order <- nsdiffs(y)

acf(diff(y, lag = 52), lag.max = 210, main = "ACF")
pacf(diff(y, lag = 52), lag.max=210, main = "PACF")



```

To further check this, we can run ndiffs() on our data set. Despite there looking like a seasonal differential, the nsdiffs() resulted in a D=0, so will test ARIMA models based off of that belief.

Looking at the ACF, there is a decay after 0 with a negative peak at lag 1, followed by a decay back to around 0. The PACF shows a significant peak at lag 1 and a slighlty less significant peak at lag 2. This leads me to believe that for our seasonal component, we should test p = 1,2 and q = 0.

```{r}
Arima1 <- Arima(y, order=c(1,1,1), seasonal=list(order=c(1,0,0), period = 52), xreg = xreg_covid, method = "CSS-ML")
Arima1$aicc

Arima2 <- Arima(y, order=c(0,1,1), seasonal=list(order=c(1,0,0), period = 52), xreg = xreg_covid, method = "CSS-ML")
Arima2$aicc

#Arima3 <- Arima(y, order=c(1,1,0), seasonal=list(order=c(1,0,0), period = 52), xreg = xreg_covid, method = "ML")
#Arima3$aicc

#Arima4 <- Arima(y, order=c(1,1,1), seasonal=list(order=c(2,0,0), period = 52), xreg = xreg_covid, method = "CSS-ML")
#Arima4$aicc

Arima5 <- Arima(y, order=c(0,1,1), seasonal=list(order=c(2,0,0), period = 52), xreg = xreg_covid, method = "CSS-ML")
Arima5$aicc

#Arima6 <- Arima(y, order=c(1,1,0), seasonal=list(order=c(2,1,0), period = 52), xreg = xreg_covid, method = "CSS-ML")
#Arima6$aicc

  
```

The models above that are commented out produced errors when attempting to run. However, Arima5 (0,1,1)(2,0,0)\[52\] produced the smallest AICc value. Now let's check the residua

```{r}
checkresiduals(Arima5)

```

Looking at the Ljung-Box test, we get a p-value of 0.08316, which is not significant. Since the aicc was only slightly smaller for the Arima2 model, I will look at that model's performance.

```{r}
checkresiduals(Arima2)


```

The p-value associated with the Ljung-Box test using Arima2 model is 0.009189, leading to it being significant and to use the Arima2 model to forecast the icnsa value.

```{r}
regression_predict_log <- predict(reg_fit2, newdata = sep_5,interval = "prediction", level = 0.95)
regression_predict <- exp(regression_predict_log)
cat("Regression Based Prediction: ", round(regression_predict[1],2), 
    " (95% PI: ", round(regression_predict[2],2), " - ", round(regression_predict[3],2), ") \n")

Arima_fc <- forecast(Arima2, h=1, xreg=matrix(0, nrow=1, ncol=1, dimnames = list(NULL, "covid")), level = 95)
Arima_forecast <- exp(as.numeric(Arima_fc$mean))
cat("ARIMA Based Prediction: ", round(Arima_forecast, 2),
    " (95% PI: ", round(exp(as.numeric(Arima_fc$lower)), 0),
    " - ", round(exp(as.numeric(Arima_fc$upper)), 0), ")\n", sep = "")
```
