```{r reprex-options, include = FALSE}
options(
  keep.source = TRUE,
  crayon.enabled = FALSE,
  reprex.current_venue = "gh"
)
```

```{r, results = 'asis', echo = FALSE, include = file.exists('.Rprofile'), eval = file.exists('.Rprofile')}
cat(sprintf("*Local `.Rprofile` detected at `%s`*", normalizePath(".Rprofile")))
```

---
title: 'Time Series HW #3'
author: "Austin Reszel"
date: "`r Sys.Date()`"
output: reprex::reprex_document
knit: reprex::reprex_render
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results='hide'}
library(ggplot2)
library(forecast)
library(fpp3)
library(tidyverse)
library(tsbox)
library(lubridate)
library(reprex)
library(zoo)
library(tseries)
library(seasonal)
library(astsa)
```

# Problem 1

## Part 1.1

First, we load in the data and examine it.

```{r}
alc_sales <- read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?id=MRTSSM4453USN",
stringsAsFactors = FALSE) 
alc_sales <- data.frame(time  = as.Date(alc_sales$observation_date),
                          value = alc_sales$MRTSSM4453USN)
alc_ts <- ts(alc_sales$value, start = c(1992, 1), frequency = 12)

autoplot(alc_ts) +
  labs(title = "Retail Sales: Beer, Wine, and Liquor Stores", x = "Time", y = "Value") +
  theme_minimal()
```

## Part 1.2

To obtain the Covid window, we will examine the difference in sales from the same month in the previous year by taking a difference using lag 12. The window in which these values are abnormally high are considered the Covid window. We find the maximum change from the previous year before 2020, and then consider the Covid window any observations larger than that maximum change.

```{r}
alc_diffs <- ts(diff(alc_sales$value, lag=12), start=c(1993,1), frequency=12)

autoplot(alc_diffs) +
  labs(title = "Retail Sales: Change from Previous Year", x = "Time", y = "Value") +
  theme_minimal()

max_pre2023 <- max(window(alc_diffs, end = c(2020, 1)))

covid_window <- alc_diffs > max_pre2023 
covid_window
idx <- which(time(covid_window) == 2024 + (7/12))
covid_window[idx] <- FALSE
```

So, the window will be considered from Feb 2020 to April 2021. We will set these observations to NA. After, we plot them and see they have been removed.

```{r}
alc_sales_nocovid <- alc_sales
alc_sales_nocovid$value[alc_sales_nocovid$time >= '2020-02-01' & alc_sales_nocovid$time <= '2021-04-01'] <- NA
alc_ts_nocovid <- ts(alc_sales_nocovid$value,start=c(1992,1), frequency=12 )
ly <- log(alc_ts_nocovid)

autoplot(alc_ts_nocovid) +
  labs(title = "Retail Sales: Covid Removed", x = "Time", y = "Value") +
  theme_minimal()
```
## Part 1.3

The structural state space model can be written as: $$T_t = \phi T_{t-1} + w_{1,t}$$ $$S_t +S_{t-1}+...+S_{t-11}=w_{t,2}$$ Then given the overall model as $y_t=T_t+S_t+\nu_t$, we can write the observation equation as $$y_t =\begin{pmatrix} 1 & 1 & 0 & 
\cdots & 0 \end{pmatrix}\begin{pmatrix} T_t \\ S_t \\ S_{t-1} \\ \cdots \\ S_{t-10} \end{pmatrix} + v_t $$ and the state equation as $$
\begin{pmatrix}
T_t \\
S_t \\
S_{t-1} \\
... \\
S_{t-10}
\end{pmatrix}
=
\begin{bmatrix}
\phi & 0 & \cdots & 0 \\
0 & -1 & \cdots & -1 \\
0 & 1 & 0 \cdots & 0 \\
\vdots & \vdots & \vdots & \vdots \\
0 \cdots & 0 & 1 & 0
\end{bmatrix}
\begin{pmatrix}
T_{t-1} \\
S_{t-1} \\
\vdots \\
S_{t-11}
\end{pmatrix}
+
\begin{pmatrix}
w_{t,1} \\
w_{t,2} \\
0 \\
\vdots \\
0
\end{pmatrix} $$

## Part 1.4

Here we fit the model.

```{r}
num <- length(ly)
A <- matrix(c(1, 1, rep(0,10)), nrow=1)
A2   <- array(rep(A, num), dim = c(1, 12, num))
A2[, , is.na(alc_ts_nocovid)] <- 0   

Linn = function(para){
 Phi = diag(0,12) 
 Phi[1,1] = para[1] 
 Phi[2,]=c(0,rep(-1,11));
 for (i in rep(3:12)){
   Phi[i,i-1] = 1
 }
 sQ1 = para[2]; sQ2 = para[3] 
 sQ  = diag(0,12); sQ[1,1]=sQ1; sQ[2,2]=sQ2
 sR = para[4] 
 kf = Kfilter(ly, A2, mu0, Sigma0, Phi, sQ, sR)
 return(kf$like)  
}

first <- ly[1:12]
m0 <- mean(first, na.rm = TRUE)
s  <- first - m0; s <- s - mean(s)
mu0    <- matrix(c(m0, s[12], s[11:2]), ncol = 1)
Sigma0 <- diag(var(ly, na.rm = TRUE), 12)
init.par <- c(0.99, 0.02, 0.02, 0.05)
est <- optim(init.par, Linn, method = "L-BFGS-B",
             lower = c(0.5, 1e-4, 1e-4, 1e-4),
             upper = c(1, Inf, Inf, Inf), hessian = TRUE)

SE  = sqrt(diag(solve(est$hessian)))
u   = cbind(estimate = est$par, SE)
rownames(u) = c("Phi11", "sigw1", "sigw2", "sigv")
u
```


```{r}
Phi = diag(0,12) 
Phi[1,1] = est$par[1] 
Phi[2,]=c(0,rep(-1,11));
for (i in rep(3:12)){
   Phi[i,i-1] = 1
 }
sQ       = diag(0,12)
sQ[1,1]  = est$par[2]
sQ[2,2]  = est$par[3]   
sR       = est$par[4]   
ks       = Ksmooth(ly, A2, mu0, Sigma0, Phi, sQ, sR)  
xs_log <- ts(ks$Xs[1,,],start=1992, freq=12) + ts(ks$Xs[2,,], start=1992, freq=12)
xs <- exp(xs_log)
```

## Part 1.5

First is a plot of the observed points and the estimates. Second is a plot of the estimates and the actual Covid data points that were previously removed.

```{r}
tsplot(alc_ts_nocovid, type = 'o', col = 4, pch = 19, main = "Observed (points) vs. Kalman-Smoothed Esimtates")
lines(xs, col = 6, lwd = 2)

tsplot(
  window(alc_ts, start = c(2020, 1), end = c(2021, 4)),
  main = "Kalman-Smoothed Estimates for Missing COVID Period",
  xlab = "Date",
  ylab = "Alcohol Sales"
)
lines(xs, col = 6, lwd = 2)
```

## Part 1.6

Now, we fit the model to account for the Covid period by incorporating an explicit intervention into the model.

```{r}
pulse_input <- as.numeric(is.na(alc_ts_nocovid))

Linn_covid = function(para){
 Phi = diag(0,12) 
 Phi[1,1] = para[1] 
 Phi[2,]=c(0,rep(-1,11));
 for (i in rep(3:12)){
   Phi[i,i-1] = 1
 }
 sQ1 = para[2]; sQ2 = para[3] 
 sQ  = diag(0,12); sQ[1,1]=sQ1; sQ[2,2]=sQ2
 sR = para[4]
 gam = para[5]
 kf = Kfilter(alc_ts, A, mu0_covid, Sigma0_covid, Phi, sQ, sR, Gam=gam, input=pulse_input)
 return(kf$like)  
}

first_covid <- alc_ts[1:12]
m0_covid <- mean(first_covid, na.rm = TRUE)
s_covid  <- first_covid - m0_covid
s_covid <- s_covid - mean(s_covid)
mu0_covid    <- matrix(c(m0_covid, s_covid[12], s_covid[11:2]), ncol = 1)
Sigma0_covid <- diag(var(alc_ts, na.rm = TRUE), 12)
init.par_covid <- c(0.99, 0.02, 0.02, 0.05, 0.05)
est_covid <- optim(init.par_covid, Linn_covid, method = "L-BFGS-B",
             lower = c(0.5, 1e-4, 1e-4, 1e-4, -Inf),
             upper = c(1, Inf, Inf, Inf, Inf), hessian = TRUE)

SE_covid  = sqrt(diag(solve(est_covid$hessian)))
u_covid   = cbind(estimate_covid = est_covid$par, SE_covid)
rownames(u_covid) = c("Phi11", "sigw1", "sigw2", "sigv", "Gam")
u_covid

```
The standard error for Gam is much larger than the actual estimates. Maybe Gamma is not needed in the model. Let's try it without Gam and see if it improves.

```{r}
Linn_covid2 = function(para){
 Phi = diag(0,12) 
 Phi[1,1] = para[1] 
 Phi[2,]=c(0,rep(-1,11));
 for (i in rep(3:12)){
   Phi[i,i-1] = 1
 }
 sQ1 = para[2]; sQ2 = para[3] 
 sQ  = diag(0,12); sQ[1,1]=sQ1; sQ[2,2]=sQ2
 sR = para[4]
 kf = Kfilter(alc_ts, A, mu0_covid, Sigma0_covid, Phi, sQ, sR)
 return(kf$like)  
}

init.par_covid2 <- c(0.99, 0.02, 0.02, 0.05)
est_covid2 <- optim(init.par_covid2, Linn_covid2, method = "L-BFGS-B",
             lower = c(0.5, 1e-4, 1e-4, 1e-4),
             upper = c(1, Inf, Inf, Inf), hessian = TRUE)

SE_covid2  = sqrt(diag(solve(est_covid2$hessian)))
u_covid2   = cbind(estimate_covid2 = est_covid2$par, SE_covid2)
rownames(u_covid2) = c("Phi11", "sigw1", "sigw2", "sigv")
u_covid2

Phi_covid = diag(0,12) 
Phi_covid[1,1] = est_covid2$par[1] 
Phi_covid[2,]=c(0,rep(-1,11));
for (i in rep(3:12)){
   Phi_covid[i,i-1] = 1
 }
sQ_covid       = diag(0,12)
sQ_covid[1,1]  = est_covid2$par[2]
sQ_covid[2,2]  = est_covid2$par[3]   
sR_covid       = est_covid2$par[4]   
ks_covid       = Ksmooth(alc_ts, A2, mu0_covid, Sigma0_covid, Phi_covid, sQ_covid, sR_covid)  
xs_covid <- ts(ks_covid$Xs[1,,],start=1992, freq=12) + ts(ks_covid$Xs[2,,], start=1992, freq=12)
```

Now, let's see how the trend differs between this and the other model we fit where the Covid observations were removed.

```{r}
xs_covid_trend <- ts(ks_covid$Xs[1,,],start=1992, freq=12)
xs_trend <- exp(ts(ks$Xs[1,,], start=1992, freq=12))

tsplot(alc_ts, type = 'o', col = 4, pch = 19, main = "Observed (points) vs. Kalman-Smoothed Trend")
lines(xs_trend, col = 1, lwd = 2)
lines(xs_covid_trend, col=2, lwd=2)
```

These appear to be very similar in terms of the extracted trend. In fact, the difference between the two models (in terms of trend) seems to be nearly negligible. So, it is possible this state-space structural model is very appropriate for handling the Covid spike, regardless of how we treat those values (missing or not).

```{r}
xs_dif <- xs_trend - xs_covid_trend
mean(xs_dif)

```

On average, the estimates for trend only differ by about 24.5, which is small given the scale is in the 1000s. We will make our estimate using the second model, which includes the Covid values. Since forcing those abnormalities to be missing doesn't produce much of a change in the model trend, we will stick with the full data set.

## Part 1.7

Now, we want to use this model to make the forecast for next month and year. First, we provide a plot. Second, we provide the estimates.

```{r}
n.ahead = 12
y       = ts(append(alc_ts, rep(0,n.ahead)), start=1992, freq=12)
rmspe   = rep(0,n.ahead) 
x00     = ks_covid$Xf[,,num]
P00     = ks_covid$Pf[,,num]
Q       = sQ%*%t(sQ) 
R       = sR%*%t(sR)
for (m in 1:n.ahead){
       xp = Phi_covid%*%x00
       Pp = Phi_covid%*%P00%*%t(Phi_covid)+Q
      sig = A%*%Pp%*%t(A)+R
        K = Pp%*%t(A)%*%(1/sig)
      x00 = xp 
      P00 = Pp-K%*%A%*%Pp
 y[num+m] = A%*%xp
 rmspe[m] = sqrt(sig) 
}


tsplot(y, type='o', main='', ylab='Alcohol Sales')
abline(v = 2026 + 6/12, lty=3) 

fc = y[(num+1):(num+n.ahead)]
window(y, start=c(2026,7), end=c(2027,7))
```

<sup>Created on `r Sys.Date()` with [reprex v`r utils::packageVersion("reprex")`](https://reprex.tidyverse.org)</sup>
