---
title: "2026 Time Series"
author: "03-regression"
format: revealjs
editor: source
toc: true
toc-depth: 1
slide-number: true
smaller: false
scrollable: true 
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
library(forecast)
library(fpp3)
library(tidyverse)
library(tsbox)
library(zoo)
library(seasonal)
library(astsa)
```

# ---- Regression Basics

## Multiple Linear Regression

Time series regression models postulate a linear relationship between the average level of a response (output) varaible $y_t$ and a bunch (say $p$) of predictor (input) variables $\textbf{x}_t = (x_{t,1}, \ldots, x_{t,p})$ for each time point $t$.

$$y_t = \beta_0 + \beta_1 x_{t,1} + \beta_2 x_{t,2} + \cdots + \beta_k x_{t,p} + \epsilon_t$$ Simplest case is 'simple linear regression' with $p = 1$. The input variables could be simple functions of time e.g. $x_{t,1} = t$ for fitting a 'Linear Trend' or $x_{t,1} = t, x_{t,2} = t^2$ for fitting a 'Quadratic Trend' or $x_{t,1} = A\cos(\theta t + \eta)$ for a periodic mean or $x_t$ could simply be other contemporaneous time series or lags of contemporaneous time series.

## Errors in the equation

Key assumption about the 'error' $\epsilon_t,\, t = 1, \ldots, T.$

-   Errors have mean zero; otherwise the fit/forecasts will be systematically biased.
-   Errors are not autocorrelated; otherwise the forecasts will be inefficient, as there is more information in the data that can be exploited.
-   Errors are unrelated to the predictor variables; otherwise there would be more information that should be included in the systematic part of the model. More generally, the input variables have to be 'exogenous', i.e. $E(\epsilon_t | x_t) = 0.$ This along with mean zero errors will mean that the imput variable is uncorrelated with the error term

## Contemporaneous input time series {.smaller}

@fig-Income shows the time series of quarterly percentage changes (growth rates) of real personal consumption expenditure, $y_t$, and real personal disposable income, $x_t$ for the US from 1970 Q1 to 2019 Q2.

```{r}
#| label: fig-Income
#| fig-cap: "Percentage changes in personal consumption expenditure and personal income for the US"
#| warning: false
us_change |>
  pivot_longer(c(Consumption, Income), names_to="Series") |>
  autoplot(value) +
  labs(y = "% change")
```

## Income versus Expenditure {.smaller}

Do you see any linear relationship between $x$ and $y$?

```{r}
us_change |>
  ggplot(aes(x = Income, y = Consumption)) +
  labs(y = "Consumption (quarterly % change)",
       x = "Income (quarterly % change)") +
  geom_point() 
```


## Income versus Expenditure {.smaller}

@fig-Scatter_Plot shows the relationship between income and expenditure.

```{r}
#| label: fig-Scatter_Plot
#| fig-cap: "Scatterplot of quarterly changes in consumption expenditure versus quarterly changes in personal income and the fitted regression line."
#| warning: false

us_change |>
  ggplot(aes(x = Income, y = Consumption)) +
  labs(y = "Consumption (quarterly % change)",
       x = "Income (quarterly % change)") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Regression of Income on Expenditure

The scatter plot shows a linear upward trend. A simple linear regression fit gives 

$$\hat{y}_t = 0.54 + 0.27 x_t $$

The fit is not very satisfactory (<span style="color:red;">how do we decide this?</span>) so need to modify the mode:  add more varaibles and/or add nonlinear terms

# ---- Lagged regression

## Regression with lagged varaibles {.smaller}


\[SS Example\~1.5\] SOI and Recruitment Correlation Analysis: @fig-SOI shows monthly values of an environmental series called the Southern Oscillation Index (SOI) and associated Recruitment (number of new fish)

```{r}
#| label: fig-SOI
#| fig-cap: "<Monthly SOI and Recruitment (estimated new fish), 1950-1987."
#| warning: false

par(mfrow = c(2,1)) # set up the graphics
plot(soi, ylab="", xlab="", main="Southern Oscillation Index")
plot(rec, ylab="", xlab="", main="Recruitment")
```

## Correlation Anaysis of SOI and Rec 

- The ACF plot shows dependence of the current value of each series on its own lagged values

```{r}
#| output-location: slide
  par(mfrow=c(1,2))
  acf(soi, 48, main="Southern Oscillation Index")
  acf(rec, 48, main="Recruitment")
```

## SOI and Rec CCF

 - In time series regression we have the opportunity to add the lagged values of the predictors as possible predictors.
- The sample cross-correlation and CCF function are defined as 

$$
\widehat{\gamma}_{xy}(h)
=
\frac{1}{n}\sum_{i=1}^{n-h}(x_{t+h} - \bar{x})(y_t - \bar{y})
$$

$$
\widehat{\rho}_{xy}(h)
= 
\frac{\widehat{\gamma}_{xy}(h)}{\sqrt{\widehat{\gamma}_x(0)\widehat{\gamma}_y(0)}}
$$


## SOI and Rec CCF

- The sample CCF in @fig-SOI2 ACf shows some departure from the cyclic component of each series and there is an obvious peak at $h = −6$. 
- This result implies that SOI measured at time $t − 6$ months is associated with the Recruitment series at time $t$.

```{r}
#| label: fig-SOI2
#| fig.cap: "The sample CCF of the two series; negative lags indicate SOI leads Recruitment"
#| warning: false
ccf(soi, rec, 48, main="SOI vs Recruitment", ylab="CCF")
```

## Lag plots for SOI  

The following figure shows scatter plots of the SOI series with different lags of SOI 

```{r}

lag1.plot(soi, 12) 
```

## Lag plots for SOI and Rec 

The following figure shows scatter plots of the SOI series with different lags of  Rec

```{r}

lag2.plot(soi, rec, 8) 
```


## Lagged regression for SOI on Rec{.scrollable}

Consider the following lagged regression,
$$R_t = \beta_0 + \beta_1 S_{t-6} + w_t$$
where $R_t$ denotes Recruitment for month $t$ and $S_{t-6}$ denotes SOI six months prior.
Assuming the $w_t$ sequence is white, the fitted model is
$$\hat{R}_t = 65.79 - 44.28_{(2.78)}S_{t-6}$$

The slope estimate is statistically significant (standard error is given in parenthesis)

```{r}
fish = ts.intersect(rec, soiL6=stats::lag(soi,-6), dframe=TRUE) 
summary(fit1 <- lm(rec~soiL6, data=fish, na.action=NULL))
```

## Fitted model for SOI-lag6 on Rec

```{r}
plot(rec ~ soiL6, data = fish)
abline(fit1)
```

# ---- Useful Predictor variables

## Trend

A linear trend can be modelled by simply using $$x_{1,t} = t$$ as a predictor,

$$y_t = \beta_0 + \beta_1 t + \epsilon_t$$


## Dummy variables

- A dummy variable is also known as an “indicator variable”.
- Example a predictor is a categorical variable taking only two values (e.g., “yes” and “no”)?

    - when forecasting daily sales and you want to take account of whether the day is a public holiday or not.
    - So the predictor takes value “yes” on a public holiday, and “no” otherwise.

-  a “dummy variable” which takes value 1 corresponding to “yes” and 0 corresponding to “no”.

 - A dummy variable can also be used to account for an outlier in the data.

 - Rather than omit the outlier, a dummy variable removes its effect.

- In this case, the dummy variable takes value 1 for that observation and 0 everywhere else.


## Seasonal Dummy variables

|          | d1, t | d2, t | d3, t | d4, t | d5, t | d6, t |
|----------|-------|-------|-------|-------|-------|-------|
| Monday   | 1     | 0     | 0     | 0     | 0     | 0     |
| Tuesday  | 0     | 1     | 0     | 0     | 0     | 0     |
| Wednesday| 0     | 0     | 1     | 0     | 0     | 0     |
| Thursday | 0     | 0     | 0     | 1     | 0     | 0     |
| Friday   | 0     | 0     | 0     | 0     | 1     | 0     |
| Saturday | 0     | 0     | 0     | 0     | 0     | 1     |
| Sunday   | 0     | 0     | 0     | 0     | 0     | 0     |
| Monday   | 1     | 0     | 0     | 0     | 0     | 0     |

## Example: Australian quarterly beer production: 

```{r, echo=FALSE}
recent_production <- aus_production |>
  filter(year(Quarter) >= 1992)
recent_production |>
  autoplot(Beer) +
  labs(y = "Megalitres",
       title = "Australian quarterly beer production")
```

## Example: Australian quarterly beer production: 

- We can model this data using a regression model with a **linear trend** and **quarterly dummy variables**,

$$y_t = \beta_0 + \beta_1 t + \beta_2 d_{2,t} + \beta_3 d_{3,t} + \beta_4 d_{4,t} + \epsilon_t,$$

- $d_{i,t} = 1$  if  $t$ is in quarter $i$ and 0 otherwise. 
- The first quarter variable has been omitted, so the coefficients associated with the other quarters are measures of the difference between those quarters and the first quarter.

## Example: Australian quarterly beer production: 

```{r}
fit_beer <- recent_production |>
  model(TSLM(Beer ~ trend() + season()))
report(fit_beer)
```

- There is an average downward trend of -0.34 megalitres per quarter. 
- On average, the second quarter has production of 34.7 megalitres lower than the first quarter
- the third quarter has production of 17.8 megalitres lower than the first quarter
- the fourth quarter has production of 72.8 megalitres higher than the first quarter.

## Example: Australian quarterly beer production: 

- Plotting fitted values

```{r}
#| output-location: slide
augment(fit_beer) |>
  ggplot(aes(x = Quarter)) +
  geom_line(aes(y = Beer, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted")) +
  scale_colour_manual(
    values = c(Data = "black", Fitted = "#D55E00")
  ) +
  labs(y = "Megalitres",
       title = "Australian quarterly beer production") +
  guides(colour = guide_legend(title = "Series"))
```

## Example: Australian quarterly beer production: 

- Plotting actual vs fitted values

```{r}
#| output-location: slide
augment(fit_beer) |>
  ggplot(aes(x = Beer, y = .fitted,
             colour = factor(quarter(Quarter)))) +
  geom_point() +
  labs(y = "Fitted", x = "Actual values",
       title = "Australian quarterly beer production") +
  geom_abline(intercept = 0, slope = 1) +
  guides(colour = guide_legend(title = "Quarter"))
```


## Intervention variables

- Often necessary to model interventions that may have affected the variable to be forecast. 
    - For example, competitor activity, advertising expenditure, industrial action


## Trading Day 

- $x_1$ = number of Mondays in month;
- $x_2$ = number of Tuesdays in month;
- $\vdots$
- $x_7$ = number of Sundays in month.


## Distributed lags

- Often useful to include advertising expenditure as a predictor. 
- The effect of advertising can last beyond the actual campaign
- Need to include lagged values of advertising expenditure

- $x_1$ = advertising for previous month;
- $x_2$ = advertising for two months previously;
- $\vdots$
- $x_m$ = advertising for $m$ months previously.


## Holiday 

Some holidays can move around in the calendar

- Easter
- Chinese New Year
- Ramadan

## Fourier series

- Alternative to using seasonal dummy variables
- Useful for long seasonal periods

If  $m$  is the seasonal period, then the first few Fourier terms are given by:

:::: {.columns}

::: {.column width="50%"}
- $x_{1,t} = \sin \left( \frac{2 \pi t}{m} \right)$
- $x_{2,t} = \cos \left( \frac{2 \pi t}{m} \right)$
- $x_{3,t} = \sin \left( \frac{4 \pi t}{m} \right)$
:::

::: {.column width="50%"}
- $x_{4,t} = \cos \left( \frac{4 \pi t}{m} \right)$
- $x_{5,t} = \sin \left( \frac{6 \pi t}{m} \right)$
- $x_{6,t} = \cos \left( \frac{6 \pi t}{m} \right)$
:::

::::


# ----- Nonlinear regression

## Nonlinear relation

```{r}
 dummy = ifelse(soi<0, 0, 1)
  fishl6 = ts.intersect(rec, soiL_l6=stats::lag(soi,-6), dummy6=stats::lag(dummy,-6), dframe=TRUE)
  attach(fishl6)
plot(soiL_l6, rec)
lines(lowess(soiL_l6, rec), col=4, lwd=2)
```



## Need for nonlinear regression {.scro}

The relationship is nonlinear and different when SOI is positive or negative. In this case, we may consider adding a dummy variable to account for this change. In particular, we fit the model:
$$
R_t = \beta_0 + \beta_1 S_{t-6} + \beta_2 D_{t-6} + \beta_3 D_{t-6} S_{t-6} + w_t
$$
where $D_t$ is a dummy variable that is 0 if $S_t < 0$ and 1 otherwise. This means that:
$$
R_t = \begin{cases} \beta_0 + \beta_1 S_{t-6} + w_t & \text{if } S_{t-6} < 0, \\ (\beta_0+\beta_2) + (\beta_1+\beta_3)S_{t-6} + w_t & \text{if } S_{t-6} \geq 0. \end{cases}$$


## Lowess fit of SOI versus lagged Rec {.scrollable}


```{r}
 dummy = ifelse(soi<0, 0, 1)
  fish = ts.intersect(rec, soiL6=stats::lag(soi,-6), dL6=stats::lag(dummy,-6), dframe=TRUE)
    summary(fit <- lm(rec~ soiL6*dL6, data=fish, na.action=NULL))
```

```{r}
attach(fish)
plot(soiL6, rec)
lines(lowess(soiL6, rec), col=4, lwd=2) 
points(soiL6, stats::fitted(fit), pch='+', col=2) 
plot(resid(fit)) # not shown ... acf(resid(fit)) # ... but obviously not noise
```

## Nonlinear Models

More generally the relationship between the input and out time series could be any nonlinear function, parametric as well as non-parametric and the input can include the current value and/or lagged valued of the input series. 

$$ y_t = g(x_t, x_{t-1}, \ldots, x_{t-H}, \theta) + \epsilon_t $$

or 

$$ y_t = g(x_t, x_{t-1}, \ldots, x_{t-H}) + \epsilon_t $$

The errors are usually taken to be additive (adding to the systematic part). The function $g$ could be any function but usually some features are important, e.g. smooth function or piecewise smooth and simple function.

